---
title: "10 Subsetting: Making big things small"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/07/21
---

# 10.1 Select specific values

```{r}
#|label: create animals vector

animals <- c("cat", "dog", "gorilla", "buffalo", "lion", "snake")
animals
```

-   Indexing: `objectp[index]`

```{r}
#|label: index one value

animals[3]
```

```{r}
#|label: index multiple values

animals[c(1, 2, 3)]
```

-   Creating a sequence: `first_number:last_number`

```{r}
#|label: create a sequence with a colon operator

1:3
```

```{r}
#|label: index with a colon operator

animals[1:3]
```

```{r}
#|label: change the order of indexes

animals[c(3, 4, 1)]
```

```{r}
#|label: remove the value at a certain index

animals[-1]
```

```{r}
#|label: remove multiple values at certin indexes using c() function

animals[-c(1, 2, 3)]
```

```{r}
#|label: results of sequences with and without parentheses

-1:3

-(1:3)
```

```{r}
#|label: remove multiple values at certin indexes using a colon operator

animals[-(1:3)]
```

# 10.2 Logical values and operations

```{r}
#|label: use booleans to return (a) certain value(s)

animals[c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE)]
```

-   `==` Equals (compared to a single value)

-   `%in%` Equals (one value match out of multiple comparisons)

-   `!=` Does not equal

-   `<` Less than

-   `>` Greater than

-   `<=` Less than or equal to

-   `>=` Greater than or equal to

```{r}
#|label: create a numbers vector

numbers <- 1:10
numbers
```

## 10.2.1 Matching a single value

```{r}
#|label: checking the equality

2 == 2
2 == 3
```

```{r}
#|label: check existence of value inside a vector

numbers == 2
numbers == 11


# Character check is case sensitive
animals == "gorilla"
animals == "Gorilla"
```

## 10.2.2 Matching multiple values

```{r}
#|label: in operator

2 %in% c(1, 2, 3)

# Character check is case sensitive
animals %in% c("cat", "dog", "gorilla")
animals %in% c("Cat", "Dog", "Gorilla")
```

## 10.2.3 Does not match

```{r}
#|label: oposite of the match

2 != 3

"cat" != "gorilla"
```

```{r}
#|label: check the match with multiple values

!animals %in% c("cat", "dog", "gorilla")
```

## 10.2.4 Greater than or less than

```{r}
#|label: check inequality

6 > 5
6 < 5
6 >= 5
5 <= 5
```

```{r}
#|label: usage with the vectors

numbers > 3
```

## 10.2.5 Combining conditional statements - or, and

-   `Condition 1 | Condition 2`

-   `Condition 1 & Condition 2`

```{r}
#|label: or (|) and and (&) operators

2 == 3 | 2 > 1

2 == 3 & 2 > 1
```

# 10.3 Subsetting a data.frame

```{r}
#|label: read data set into R

ucr <- readRDS("data/offenses_known_yearly_1960_2020.rds")
```

```{r}
#|label: first 6 rows and columns

ucr[1:6, 1:6]
```

-   Number of rows

-   Number of columns

```{r}
#|labels: number of rows and columns

nrow(ucr)

ncol(ucr)
```

```{r}
#|label: the name of each variable

names(ucr)
```

-   Subsetting goal for this data set:

    -   Only rows where the state equals Colorado

    -   Only rows where the year is 2011-2017

    -   Only the following columns: *actual_murder*, *state*, *year*, *population*, *ori*, *agency_name*

## 10.3.1 Select specific columns

-   `data$column`

```{r}
#|label: the first 6 rows of the column agency_name

head(ucr$agency_name)
```

```{r}
#|label: the first 6 rows of the column actual murder

head(ucr$actual_murder)
```

## 10.3.2 Select specific rows

-   `[row, column]`

```{r}
#|label: the first row of the first column

ucr[1, 1]
```

```{r}
#|label: error

# ucr[1, ucr$agency_name]
# Error: unexpected symbol in "ucr[1 ucr"

# ucr[1 ucr$agency_name]
# Error: unexpected symbol in "ucr[1 ucr"
```

```{r}
#|label: multiple rows and multiple columns

ucr[1:6, 1:6]
```

```{r}
#|label: a vector of the names of the columns

ucr[1:6, c("ori", "year")]
```

```{r}
#|label: the first row of the every column

ucr[1, ]
```

```{r}
#|label: every row of the first column

ucr[ , 1]
```

```{r}
#|label: the below code returns the column that matches the numner inside the square brackets

head(ucr[1])
```

```{r}
#|label: asking for any number above the number of the columns will returns an error

# head(ucr[1000])
# Error in `[.data.frame`(ucr, 1000) : undefined columns selected
```

```{r}
#|label: with dollar sign notation there is no need to indicate any column, only row or rows must be specified

ucr$agency_name[15]
```

## 10.3.3 Subset Colorado data

-   Subsetting goal for this data set:

    -   Only rows where the state equals Colorado

    -   Only rows where the year is 2011-2017

    -   Only the following columns: actual_murder, state, year, population, ori, agency_name

    ### 10.3.3.1 Subsetting using base R

```{r}
#|label: Base R - Step 1. Only rows where the state equals Colorado

colorado <- ucr[ucr$state == "colorado", ]
```

```{r}
#|label: Base R - Step 2. Only rows where the year is 2011-2017

colorado <- colorado[colorado$year %in% 2011:2017, ]
```

```{r}
#|label: Base R - Step 3. Only the following columns - actual_murder, state, year, population, ori, agency_name


colorado <- colorado[ , c("actual_murder",
                          "state",
                          "year",
                          "population",
                          "ori",
                          "agency_name")]
```

```{r}
#|label: check the results of subsetting - Base R

unique(colorado$state)

unique(colorado$year)
```

### 10.3.3.2 Subsetting using `dplyr`

```{r}
#|label: load the dplyr package

library(dplyr)
```

```{r}
#|label: dplyr - Step 1. Only rows where the state equals Colorado

colorado <- filter(ucr, state == "colorado")
```

```{r}
#|label: dplyr - Step 2. Only rows where the year is 2011-2017

colorado <- filter(ucr, year %in% 2011:2017)
```

```{r}
#|label: dplyr - Step 1 and Step 2 all together

colorado <- filter(ucr, state == "colorado", year %in% 2011:2017)
```

```{r}
#|label: dplyr - Step 3. Only the following columns - actual_murder, state, year, population, ori, agency_name

colorado <- select(colorado, actual_murder, state, year, population, ori, agency_name)
```

```{r}
#|label: check the results of subsetting - dplyr

unique(colorado$state)

unique(colorado$year)
```
