---
title: "Graphing with ggplot2"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/08/07
---

```{r}
#|label: load the ggplot2 package

library(ggplot2)
```

```{r}
#|label: load the data

load("data/apparent_per_capita_alcohol_consumption.rda")
```

```{r}
#|label: change the name of the data frame

alcohol <- apparent_per_capita_alcohol_consumption
```

```{r}
#|label: filter the data to only Pennsylvania

penn_alcohol <- alcohol[alcohol$state == "pennsylvania", ]
```

## 14.1 What does the data look like?

```{r}
#|label: check the heading of the data set

head(penn_alcohol)
```

-   So each row of the data is a single year of data for Pennsylvania. It includes alcohol consumption for wine, liquor, beer, and total drinks - both as gallons of ethanol (a hard unit to interpret) and more traditional measures such as glasses of wine or number of beers.

-   The original data only included the gallons of ethanol data, which were converted to the more understandable units.

-   If you encounter data with odd units, it is a good idea to convert it to something easier to understand - especially if you intend to show someone else the data or results.

## 14.2 Graphing data

For `ggplot()` we need to specify four things:

1.  The data set

2.  The x-axis variable

3.  The y-axis variable

4.  The type of graph - e.g. line, point, etc.

Some useful types of graphs are:

-   `geom_point()` - A point graph, can be used for scatter plots

-   `geom_line()` - A line graph

-   `geom_bar()` - A barplot

-   `geom_smooth()` - Adds a regression line to the graph

## 14.3 Time-series plots

```{r}
#|label: add data set as well as x and y axis

ggplot(penn_alcohol, aes(x = year,
                        y = number_of_beers))
```

```{r}
#|label: convert year column from character to numeric

penn_alcohol$year <- as.numeric(penn_alcohol$year)
```

```{r}
#|label: add data set as well as x and y axis again

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_beers))
```

```{r}
#|label: add data set, x and y axis and graph type - scatter plot

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_beers)) +
  geom_point()
```

```{r}
#|label: add data set, x and y axis and graph type - line graph

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_beers)) +
  geom_line()
```

```{r}
#|label: add data set, x and y axis and graph type - scatter plot and line graph

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_beers)) +
  geom_point() +
  geom_line()
```

```{r}
#|label: force the y-axis to begin at 0

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_beers)) +
  geom_point() +
  geom_line() +
  expand_limits(y = 0)
```

```{r}
#|label: adjust color and size of the line

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_beers)) +
  geom_line(color = "forestgreen", linewidth = 1.3)
```

```{r}
#|label: add graph title, x-axis label and y-axis label

ggplot(penn_alcohol, aes(x = year,
                         number_of_beers)) +
  geom_line(color = "forestgreen", linewidth = 1.3) +
  xlab("Year") +
  ylab("Number of Beers") +
  ggtitle("PA Annual Beer Consumtion Per Capita (1977-2017)")
```

```{r}
#|label: time-series plot for wine consumption

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_glasses_wine)) +
  geom_line()
```

```{r}
#|label: time series plots for wine and liquor consumption

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_glasses_wine)) +
  geom_line() +
  geom_line(aes(x = year,
                y = number_of_shots_liquor))
```

```{r}
#|label: time series plots for wine and liquor consumption - different colors

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_glasses_wine,
                         color = "Glasses of Wine")) +
  geom_line() +
  geom_line(aes(x = year,
                y = number_of_shots_liquor,
                color = "Shots of Liquor")) 
```

```{r}
#|{r}
#|label: time series plots for wine and liquor consumption - different colors, more descriptive legend

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_glasses_wine,
                         color = "Glasses of Wine")) +
  geom_line() +
  geom_line(aes(x = year,
                y = number_of_shots_liquor,
                color = "Shots of Liquor")) +
  labs(color = "Alcohol Type")
```

```{r}
#|label: time series plots for wine and liquor consumption - different colors, more descriptive legend, legend moved to the bottom

ggplot(penn_alcohol, aes(x = year,
                         y = number_of_glasses_wine,
                         color = "Glasses of Wine")) +
  geom_line() +
  geom_line(aes(x = year,
                y = number_of_shots_liquor,
                color = "Shots of Liquor")) +
  labs(color = "Alcohol Type") +
  theme(legend.position = "bottom")
```

## 14.4 Scatter plots

```{r}
#|label: comparison of liquor to beer sales

ggplot(penn_alcohol, aes(x = number_of_shots_liquor,
                         y = number_of_beers)) +
  geom_point()
```

## 14.5 Color blindness

A helpful site for choosing colors for graphs and maps is [Color Brewer.](http://colorbrewer2.org/)

```{r}
#|label: colorblind safe colors for the line graphs

ggplot(penn_alcohol, aes(x = year, 
                         y = number_of_glasses_wine,
                         color = "Glasses of Wine")) +
  geom_line() +
  geom_line(aes(x = year,
                y = number_of_shots_liquor,
                color = "Shots of Liquor")) +
  labs(color = "Alcohol Type") +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#7570b3", "#d95f02"))
```
