---
title: "Exploratory data analysis"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/07/24
---

When you first start working on new data it is important to spend some time getting familiar with the data. This includes understanding how many rows and columns it has, what each row means (is each row an offender? a victim? crime in a city over a day/month/year?, etc.), and what columns it has. **Basically you want to know if the data is capable of answering the question you are asking.**

While not a comprehensive list, the following is a good start for exploratory data analysis of new data sets.

-   What are the units (what does each row represent?)?

-   What variables are available?

-   What time period does it cover?

-   Are there outliers? How many?

-   Are there missing values? How many?

```{r}
#|label: load the data set

load("data/ucr2017.rda")
```

```{r}
#|label: check the first 6 rows

head(ucr2017)
```

```{r}
#|label: check column names

names(ucr2017)
```

# 11.1 Summary and Table

```{r}
#|label: summary statistics of numeric variables

summary(ucr2017)
```

```{r}
#|label: each unique value and its frequency in a vector and/or (column of the data frame)


table(ucr2017$year)

# Using it with numeric variables to show the frequency of each numeric value
head(table(ucr2017$population))
```

```{r}
#|label: use head for a single vector or a column of a data frame

head(ucr2017$agency_name)
```

```{r}
#|label: real use case of table() function where the number of categories is important, such as years, states, neighbourhoods etc.

# The default is alphabetical order

table(ucr2017$state)
```

```{r}
#|label: sort in increasing order - default to sort() function

sort(table(ucr2017$state))
```

```{r}
#|label: sort in decreasing order

sort(table(ucr2017$state), decreasing = TRUE)
```

# 11.2 Graphing

```{r}
#|label: relationship between actual robbery total and actual murder per year

plot(ucr2017$actual_murder,
     ucr2017$actual_robbery_total)
```

```{r}
#|label: load the dplyr package

library(dplyr)
```

```{r}
#|label: filter the cities with a population over 1000000

ucr2017_big_cities <- filter(ucr2017,
                             population > 1000000)
```

```{r}
#|label: graphing the relationship between actual robbery total and actual murder per year in the new data

plot(ucr2017_big_cities$actual_murder,
     ucr2017_big_cities$actual_robbery_total)
```

```{r}
#|label: add labels to the scatterplot

plot(ucr2017_big_cities$actual_murder,
     ucr2017_big_cities$actual_robbery_total,
     xlab = "Murders",
     ylab = "Robberies",
     main = "Relationship between murder and robbery")
```

# 11.3 Aggregating (summaries of groups)

-   Aggregating data means that we group values at some higher level than they currently are (e.g. from agency to state, from day to month, from city street to city neighborhood) and then do some mathematical operation of our choosing (in our case usually sum) to that group.

```{r}
#|label: read the data to R

ucr <- readRDS("data/offenses_known_yearly_1960_2020.rds")
```

```{r}
#|label: filter rows and select relevant columns

colorado <- filter(ucr, state == "colorado",
                   year %in% 2011:2017)

colorado <- select(colorado, 
                   actual_murder, 
                   actual_robbery_total,
                   state,
                   year,
                   population,
                   ori,
                   agency_name)
```

```{r}
#|label: group the data

colorado <- group_by(colorado, year)
```

```{r}
#|label: summarize the data - aggregate a single column

summarize(colorado, sum(actual_murder))
```

```{r}
#|label: summarize the data - aggregate another column

summarize(colorado, sum(actual_murder),
          sum(actual_robbery_total))
```

```{r}
#|label: summatize the data - use different mathematical operation

summarize(colorado, sum(actual_murder),
          sum(actual_robbery_total),
          mean(actual_robbery_total))
```

```{r}
#|label: create aggregated data set with appropriate column names

colorado_agg <- summarize(colorado,
                          murders = sum(actual_murder),
                          robberies = sum(actual_robbery_total),
                          population = sum(population))

colorado_agg
```

-   Standardize the murder data by creating **a murder and robbery rate per 100,000 people**

```{r}
#|label: add a murder rate column

colorado_agg$murder_rate <- colorado_agg$murders / colorado_agg$population * 100000
```

```{r}
#|label: add a robbery rate column

colorado_agg$robbery_rate <- colorado_agg$robberies / colorado_agg$population * 100000
```

```{r}
#|label: add new columns in one go

mutate(colorado_agg, 
       murder_rate = murders / population * 100000,
       robbery_rate = robberies / population * 10000)
```

```{r}
#|label: creating time-series graph - year vs murder rate - default is scatter plot for plot() function

plot(x = colorado_agg$year,
     y = colorado_agg$murder_rate)
```

```{r}
#|label: creating time-series graph - year vs murder rate - line graph

plot(x = colorado_agg$year,
     y = colorado_agg$murder_rate,
     type = "l")
```

```{r}
#|label: add labels to the line graph

plot(x = colorado_agg$year,
     y = colorado_agg$murder_rate,
     type = "l",
     xlab = "Year",
     ylab = "Murders per 100k Population",
     main = "Murder Rate in Colorado, 2011-2017")
```

# 11.4 Pipes in `dplyr`

```{r}
#|label: without pipe

colorado <- filter(ucr, state == "colorado", 
                   year %in% 2011:2017)

colorado <- select(colorado,
                   actual_murder,
                   actual_robbery_total,
                   state,
                   year,
                   population,
                   ori,
                   agency_name)

head(colorado)
```

```{r}
#| label: with pipe

colorado <- ucr |> filter(state == "colorado",
                          year %in% 2011:2017) |> 
  select(actual_murder,
         actual_robbery_total,
         state, year, population,
         ori, agency_name)

head(colorado)
```

```{r}
#|label: write code using pipe in conventional way - have a new line after the pipe and after each comma

colorado <- ucr |> 
  filter(state == "colorado",
         year %in% 2011:2017) |> 
  select(actual_murder,
         actual_robbery_total,
         state,
         year,
         population,
         ori,
         agency_name)
```
