---
title: "Regular Expressions"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/08/04
---

-   `grep()` - Find

-   `gsub()` - Find and Replace

```{r}
#|label: crime categories vector for practice

crimes <- c(
  "Arson",
  "Assault",                                  
  "Burglary",                                 
  "Case Closure",                             
  "Civil Sidewalks",                          
  "Courtesy Report",                          
  "Disorderly Conduct",                       
  "Drug Offense",                             
  "Drug Violation",                           
  "Embezzlement",                             
  "Family Offense",                           
  "Fire Report",                              
  "Forgery And Counterfeiting",               
  "Fraud",                                    
  "Gambling",                                 
  "Homicide",                                 
  "Human Trafficking (A), Commercial Sex Acts",
  "Human Trafficking, Commercial Sex Acts",   
  "Juvenile Offenses",                        
  "Larceny Theft",                            
  "Liquor Laws",                              
  "Lost Property",                            
  "Malicious Mischief",                       
  "Miscellaneous Investigation",              
  "Missing Person",                           
  "Motor Vehicle Theft",                      
  "Motor Vehicle Theft?",                     
  "Non-Criminal",                             
  "Offences Against The Family And Children", 
  "Other",                                    
  "Other Miscellaneous",                      
  "Other Offenses",                           
  "Prostitution",                             
  "Rape",                                     
  "Recovered Vehicle",                        
  "Robbery",                                  
  "Sex Offense",                              
  "Stolen Property",                          
  "Suicide",                                  
  "Suspicious",                               
  "Suspicious Occ",                           
  "Traffic Collision",                        
  "Traffic Violation Arrest",                 
  "Vandalism",                                
  "Vehicle Impounded",                        
  "Vehicle Misplaced",                        
  "Warrant",                                  
  "Weapons Carrying Etc",                     
  "Weapons Offence",                          
  "Weapons Offense"
)
```

## 12.1 Finding patterns in text with `grep()`

-   The syntax for `grep()` is:

    -   `grep("pattern", data)`

-   where pattern is the pattern you are searching for, such as “a” if you want to find all values with the letter a.

```{r}
#|label: find a pattern "a" in crimes vector

grep("a", crimes)
```

```{r}
#|label: subsetting the crimes vector - crime categories containing the pattern "a"

crimes[grep("a", crimes)]
```

```{r}
#|label: find a pattern "Theft" in a crime vector

grep("Theft", crimes)
```

```{r}
#|label: subsetting the crime vector - crime categories containing the pattern "Theft"

crimes[grep("Theft", crimes)]
```

```{r}
#|label: display values of the pattern

grep("Theft", crimes, value = TRUE)
```

```{r}
#|label: grep() is case sensitive

grep("theft", crimes, value = TRUE)
```

```{r}
#|label: ignoring the case sensitivity in grep()

grep("theft", crimes, value = TRUE, ignore.case = TRUE)
```

```{r}
#|label: inverting the result of the pattern

grep("theft", crimes, value = TRUE, ignore.case = TRUE, invert = TRUE)
```

## 12.2 Finding and replacing patterns in text with `gsub()`

-   The syntax of `gsub()` is similar to `grep()` with the addition of a pattern to replace the pattern we found.

    -   `gsub("find_pattern", "replace_pattern", data)`

```{r}
#|label: replacing a pattern "a" with a pattern "z"

gsub("a", "z", "cat")
```

```{r}
#|label: gsub() is case sensitive

gsub("A", "z", "cat")
```

```{r}
#|label: ignoring case sensitivity in gsub()

gsub("A", "z", "cat", ignore.case = TRUE)
```

```{r}
#|label: replaceing a pattern "Offence" with a pattern "Offense"

gsub("Offence", "Offense", crimes)
```

```{r}
#|label: replacing a pattern "a" with a pattern "" - replacing with an empty pattern

gsub("a", "", crimes)
```

## 12.3 Useful special characters

### 12.3.1 Multiple characters

```{r}
#|label: searching for multiple characters

grep("[xyz]", crimes, value = TRUE)
```

```{r}
#|label: the order does not matter

grep("[zyx]", crimes, value = TRUE)
```

```{r}
#|label: it also works for numbers - there is no data in crimes vector

grep("[0123456789]", crimes, value = TRUE)
```

```{r}
#|label: search for a repeated pattern - vowels

grep("[aeiou][aeiou][aeiou]", crimes, value = TRUE)
```

```{r}
#|label: making intervals

grep("[x-z]", crimes, value = TRUE)
```

### 12.3.2 n-many of previous character `{n}`

-   `{n}` means the preceding item will be matched exactly n times.

```{r}
#|label: n-number of mathces - searching for the three vowels pattern

grep("[aeiou]{3}", crimes, value = TRUE)
```

### 12.3.3 n-many to m-many of previous character `{n,m}`

-   While `{n}` says “the previous character (or characters inside a `[]`) must be present exactly n times”, we can allow a range by using `{n,m}`.

```{r}
#|label: n to m number of mathces - searching for 2-3 vowels in a pattern

grep("[aeiou]{2,3}", crimes, value = TRUE)
```

```{r}
#|label: searching for a pattern with exactly m times match - exactly 3 vowels in a row

grep("[aeiou]{3,3}", crimes, value = TRUE)

# Equivalent
print("############")

grep("[aeiou]{3}", crimes, value = TRUE)
```

```{r}
#|label: up to m times including zero times - up to 3 vowels

grep("[aeiou]{,3}", crimes, value = TRUE)
```

```{r}
#|label: at least n times

grep("[aeiou]{3,}", crimes, value = TRUE)
```

### 12.3.4 Start of string

-   The `^` symbol (called a caret) signifies that what follows it is the **start** of the string.

```{r}
#|label: search for a starting pattern

grep("^Family", crimes, value = TRUE)
```

### 12.3.5 End of string `$`

-   The dollar sign `$` acts similar to the caret `^` except that it signifies that the value before it is the **end** of the string. 

```{r}
#|label: search for a ending pattern

grep("Theft$", crimes, value = TRUE)
```

### 12.3.6 Anything `.`

-   The `.` symbol is a stand-in for any value. 

```{r}
#|label: search for a pattern that contains "anything"

grep("Weapons Offen.e", crimes, value = TRUE)
```

### 12.3.7 One or more of previous `+`

-   The `+` means that the character immediately before it is present at least one time.

-   This is the same as writing `{1,}`.

-   If we wanted to find all values with only two words, we would start with some number of letters followed by a space followed by some more letters and the string would end.

```{r}
#|label: values with only two words

grep("^[A-Za-z]+ [A-Za-z]+$", crimes, value = TRUE)
```

```{r}
#|label: equivalent

grep("^[A-Za-z]{1,} {1,}[A-Za-z]{1,}$", crimes, value = TRUE)
```

### 12.3.8 Zero or more of previous `*`

-   The `*` special character says match zero or more of the previous character and is the same as `{0,}`.

-   Combining `.` with `*` is powerful when used in `gsub()` to delete text before or after a pattern.

```{r}
#|label: a pattern that searches the text for the word “Weapons” and then deletes any text after that

gsub("Weapons.*", "Weapons", crimes)
```

### 12.3.9 Multiple patterns `|`

-   The vertical bar `|` special character allows us to check for multiple patterns. It essentially functions as “pattern A or Pattern B” with the `|` symbol replacing the word “or” (and making sure to not have any space between patterns.).

```{r}
#|label: check the crimes for the word “Drug” or the word “Weapons”

grep("Drug|Weapons", crimes, value = TRUE)
```

### 12.3.10 Parentheses `()`

-   Parentheses act similar to the square brackets `[]`, where we want everything inside but with parentheses the values must be in the proper order.

```{r}
#|label: values with a word "Offense"

grep("(Offense)", crimes, value = TRUE)
```

```{r}
#|label: search for the values with "Drug or Weapon Offense"

grep("(Drug|Weapons) Offense", crimes, value = TRUE)
```

### 12.3.11 Optional text `?`

-   The question mark indicates that the character immediately before the `?` is optional.

```{r}
#|label: search for the term "offens"

grep("Offens?", crimes, value = TRUE)
```

```{r}
#|label: combining ? with () and | to get both spellings of Weapon Offense

grep("(Drug|Weapons) Offens?", crimes, value = TRUE)
```

## 12.4 Changing capitalization

```{r}
#|label: make all the values capitalized

toupper(crimes)
```

```{r}
#|label: make each value lower case

tolower(crimes)
```
