---
title: "More graphing with ggplot2"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/08/10
---

```{r}
#|label: load the readr package

library(readr)
```

```{r}
#|label: read the csv file into R and convert it from tibble to data frame

shootings <- read_csv("data/fatal-police-shootings-data.csv")

shootings <- as.data.frame(shootings)
```

## 15.1 Exploring data

```{r}
#|label: number of columns and rows

ncol(shootings)

nrow(shootings)
```

```{r}
#|label: first 6 rows of the data set

head(shootings)
```

```{r}
#|label: summary of the date column

summary(shootings$date)
```

```{r}
#|label: frequency of the causes of death

table(shootings$manner_of_death)
```

```{r}
#|label: percentage division of the causes of death

table(shootings$manner_of_death) / nrow(shootings) * 100
```

```{r}
#|label: armed status of the victims

table(shootings$armed) / nrow(shootings) * 100
```

```{r}
#|label: sorting the results in decreasing decreasing order

# get percentages of the arms the victims had
temp <- table(shootings$armed) / nrow(shootings) * 100

# sort the result in a decreasing order
temp <- sort(temp, decreasing = TRUE)

# round the results to have 2 decimal points
temp <- round(temp, digits = 2)
```

```{r}
#|label: genders of the victims

table(shootings$gender) / nrow(shootings) * 100
```

```{r}
#|label: race of the victims

temp <- table(shootings$race) / nrow(shootings) * 100
temp <- sort(temp)
temp <- round(temp, digits = 2)
temp
```

## 15.2 Graphing a single numeric variable

```{r}
#|label: load the ggplot2 package

library(ggplot2)
# Warning: package 'ggplot2' was built under R version 4.2.2
```

```{r}
#|label: add data and age variable

ggplot(shootings, aes(x = age))
```

### 15.2.1 Histogram

-   The histogram is a very common type of graph for a single numeric variable. Histograms group a numeric variable into categories and then plot them, with the heights of each bar indicating how common the group is.

```{r}
#|label: histogram to visualize age variable - default bin parameter set to 30

ggplot(shootings, aes(x = age)) +
  geom_histogram()
```

```{r}
#|label: set bin parameter to 15

ggplot(shootings, aes(x = age)) +
  geom_histogram(bins = 15)
```

```{r}
#|label: set bin parameter to 45

ggplot(shootings, aes(x = age)) +
  geom_histogram(bins = 45)
```

```{r}
#|label: display proportions instead of the number

ggplot(shootings, aes(x = age)) +
  geom_histogram(
    aes(y = after_stat(count) / after_stat(sum(count))))
```

### 15.2.2 Density plot

-   Density plots are essentially smoothed versions of histograms.

-   They’re especially useful for numeric variables, which are not integers (integers are whole numbers).

-   They’re also useful when you want to be more precise than a histogram as they are - to simplify - histograms where each bar is very narrow.

-   Note that the y-axis of a density plot is automatically labeled “density” and has very small numbers.

-   Interpreting the y-axis is fairly hard to explain to someone not familiar with statistics so I’d caution against using this graph unless your audience is already familiar with it

```{r}
#|label: density plot

ggplot(shootings, aes(x = age)) +
  geom_density()
```

### 15.2.3 Count graph

-   A count graph is essentially a histogram with a bar for every value in the numeric variable - like a less-smooth density plot.

-   Note that this won’t work well if you have too many unique values so I’d strongly recommend rounding the data to the nearest whole number first if you don’t already have an integer.

-   Our age variable is already rounded so we don’t need to do that. To make a count graph, we add `stat_count()` to the `ggplot()`.

```{r}
#|label: count graph - a single bar for every age in the data

ggplot(shootings, aes(x = age)) +
  stat_count()
```

```{r}
#|label: count graph - proportions

ggplot(shootings, aes(x = age)) +
  stat_count(aes(y = after_stat(count) / after_stat(sum(count))))
```

## 15.3 Graphing a categorical variable
