---
title: "A soup to nuts project example"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/06/20
---

# 1.1 Big picture data example

```{r}
#|label: Percent change in murders in 2020 relative to 2019: Vector and data set manipulation up to the graphing

# Loading packages
library(dplyr)       # Used to aggregate data
library(ggplot2)     # Used to make the graph
library(crimeutils)  # Used to capitalize words in a column
library(tidyr)       # Used to reshape the data


# Load in the data
shr <- readRDS("data/shr_1976_2020.rds")

# See which agencies reported in 2019 and 2020
# An "ori" is a unique identifier code for agencies in FBI data
agencies_2019 <- shr$ori[shr$year == 2019]
agencies_2020 <- shr$ori[shr$year == 2020]


# Get which agencies reported in both years so we have an
# apples-to-apples comparison
agencies_in_both <- agencies_2019[agencies_2019 %in% agencies_2020]


# Keep just data from 2019 and 2020 and where the agencies
# is one of the agencies chosen above. Also keep only murder and 
# nonnegligent manslaughter (so excluding  negligent manslaughter).
shr_2019_2020 <- shr[shr$year %in% 2019:2020, ]
shr_2019_2020 <- shr_2019_2020[shr_2019_2020$ori %in% agencies_in_both, ]
shr_2019_2020 <- shr_2019_2020[shr_2019_2020$homicide_type %in% 
                                 "murder and nonnegligent manslaughter", ]


# Get the number of murders by victim-offender relationship in 2019 and 2020
# Then find the percent change in murders by this group from 2019 to 2020
# Sort data by smallest to largest percent change
shr_difference_victim_offender <- 
  shr_2019_2020 |> 
  group_by(year) |> 
  count(victim_1_relation_to_offender_1) |>
  spread(year, n) |> 
  mutate(difference = `2020` - `2019`, 
         percent_change = difference / `2019` * 100,
         victim_1_relation_to_offender_1 = 
           capitalize_words(victim_1_relation_to_offender_1)) |> 
  filter(`2019` >= 50) |> 
  arrange(percent_change)

shr_difference_victim_offender


# This is only for the graph. By default graphs order alphabetically
# but this makes sure it orders it based on the ordering we made above
# (smallest to largest percent change)

shr_difference_victim_offender$victim_1_relation_to_offender_1 <- 
  factor(shr_difference_victim_offender$victim_1_relation_to_offender_1, 
         levels = shr_difference_victim_offender$victim_1_relation_to_offender_1)
```

```{r}
#|label: Percent change in murders in 2020 relative to 2019: Graph


# Makes a barplot showing the percent change from 2019 to 2020 in number
# of murders by victim group. Labels the x-axis and the y-axis, shifts 
# the graph so that relationship labels are on the y-axis for easy reading. 
# And finally uses the "crim" theme that changes the colors in the graph to
# make it a little easier to see.
ggplot(shr_difference_victim_offender, aes(x = victim_1_relation_to_offender_1,
                           y = percent_change)) + 
  geom_bar(stat = "identity") +
  ylab("% Change, 2020 Vs. 2019") +
  xlab("Victim Relative to Murderer") +
  coord_flip() +
  theme_crim()
```

# 1.2 Little picture data example

1.  Load the packages used

2.  Load the data

3.  Clean the data

4.  Aggregate the data

5.  Make the graph

## 1.1.1 Loading packages

```{r}
#|label: Loading packages

library(dplyr)       # Used to aggregate data
library(ggplot2)     # Used to make the graph
library(crimeutils)  # Used to capitalize words in a column
library(tidyr)       # Used to reshape the data
```

## 1.2.2 Loading data

-   SHR - FBI's Supplementary Homicide Report

-   ORI - Originating Agency Identifier

```{r}
#|label: Loading data

# Load in the data
shr <- readRDS("data/shr_1976_2020.rds")
```

## 1.2.3 Cleaning

-   Non-negligent manslaughter, as defined by the FBI's Uniform Crime Reporting (UCR) Program, is the willful (non-negligent) killing of one human being by another. This means it involves a conscious decision to cause death, without the element of negligence or accident. It's crucial to distinguish this from negligent manslaughter, which involves death caused by recklessness or gross negligence, and from murder, which requires intent to kill.

-   Cleaning steps:

    1.  Create a vector of agencies which reported in 2019.

    2.  Create a vector of agencies which reported in 2020.

    3.  Create a vector agencies which reported in both 2019 and 2020.

    4.  Subset the data to only data from 2019 and 2020 and where the agency reported in both years.

    5.  Subset the data to include only murder and non-negligent manslaughter.

```{r}
#|label: Cleaning

# See which agencies reported in 2019 and 2020
# An "ori" is a unique identifier code for agencies in FBI data
agencies_2019 <- shr$ori[shr$year == 2019]
agencies_2020 <- shr$ori[shr$year == 2020]


# Get which agencies reported in both years so we have an
# apples-to-apples comparison
agencies_in_both <- agencies_2019[agencies_2019 %in% agencies_2020]


# Keep just data from 2019 and 2020 and where the agencies
# is one of the agencies chosen above. Also keep only murder and 
# nonnegligent manslaughter (so excluding  negligent manslaughter).
shr_2019_2020 <- shr[shr$year %in% 2019:2020, ]
shr_2019_2020 <- shr_2019_2020[shr_2019_2020$ori %in% agencies_in_both, ]
shr_2019_2020 <- shr_2019_2020[shr_2019_2020$homicide_type %in% 
                                 "murder and nonnegligent manslaughter", ]
```

## 1.2.4 Aggregating

-   Aggregation steps:

    1.  Aggregate the data to year level to see how many murders there were for each victim-offender relationship group.

    2.  Find the difference between the number of murder in 2019 and 2020 and calculate percent difference.

    3.  Remove any rows there were fewer than 50 murders of that victim-offender relationship type in 2019.

    4.  Arrange the data from smallest to largest difference.

```{r}
#|label: Aggregating

# Get the number of murders by victim-offender relationship in 2019 and 2020
# Then find the percent change in murders by this group from 2019 to 2020
# Sort data by smallest to largest percent change
shr_difference_victim_offender <- 
  shr_2019_2020 |> 
  group_by(year) |> 
  count(victim_1_relation_to_offender_1) |>
  spread(year, n) |> 
  mutate(difference = `2020` - `2019`, 
         percent_change = difference / `2019` * 100,
         victim_1_relation_to_offender_1 = 
           capitalize_words(victim_1_relation_to_offender_1)) |> 
  filter(`2019` >= 50) |> 
  arrange(percent_change)

shr_difference_victim_offender
```

## 1.2.5 Graphing

-   Graphing steps:

    1.  Include the data set and columns to graph.

    2.  Tell what type of graph to create in this case a bar chart.

    3.  Set x-axis and y-axis labels.

    4.  Include the lines of code that affect how the graph looks.

```{r}
#|label: Graphing


# This is only for the graph. By default graphs order alphabetically
# but this makes sure it orders it based on the ordering we made above
# (smallest to largest percent change)

shr_difference_victim_offender$victim_1_relation_to_offender_1 <- 
  factor(shr_difference_victim_offender$victim_1_relation_to_offender_1, 
         levels = shr_difference_victim_offender$victim_1_relation_to_offender_1)


# Makes a barplot showing the percent change from 2019 to 2020 in number
# of murders by victim group. Labels the x-axis and the y-axis, shifts 
# the graph so that relationship labels are on the y-axis for easy reading. 
# And finally uses the "crim" theme that changes the colors in the graph to
# make it a little easier to see.
ggplot(shr_difference_victim_offender, aes(x = victim_1_relation_to_offender_1,
                           y = percent_change)) + 
  geom_bar(stat = "identity") +
  ylab("% Change, 2020 Vs. 2019") +
  xlab("Victim Relative to Murderer") +
  coord_flip() +
  theme_crim()
```

# 1.3 Reusing and modifying code

```{r}
#|label: Reusing and modifying code


shr_difference_offender_weapon <- 
  shr_2019_2020 |> 
  group_by(year) |> 
  count(offender_1_weapon) |>
  spread(year, n) |> 
  mutate(difference = `2020` - `2019`, 
         percent_change = difference / `2019` * 100,
         offender_1_weapon = 
           capitalize_words(offender_1_weapon)) |> 
  filter(`2019` >= 50) |> 
  arrange(percent_change)

shr_difference_offender_weapon


shr_difference_offender_weapon$offender_1_weapon <- 
  factor(shr_difference_offender_weapon$offender_1_weapon , 
         levels = shr_difference_offender_weapon$offender_1_weapon)

ggplot(shr_difference_offender_weapon, aes(x = offender_1_weapon,
                                           y = percent_change)) + 
  geom_bar(stat = "identity") +
  ylab("% Change, 2020 Vs. 2019") +
  xlab("Offender Weapon") +
  coord_flip() +
  theme_crim()
```
