---
title: "More graphing with ggplot2"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/08/10
---

```{r}
#|label: load the readr package

library(readr)
```

```{r}
#|label: read the csv file into R and convert it from tibble to data frame

shootings <- read_csv("data/fatal-police-shootings-data.csv")

shootings <- as.data.frame(shootings)
```

## 15.1 Exploring data

```{r}
#|label: number of columns and rows

ncol(shootings)

nrow(shootings)
```

```{r}
#|label: first 6 rows of the data set

head(shootings)
```

```{r}
#|label: summary of the date column

summary(shootings$date)
```

```{r}
#|label: frequency of the causes of death

table(shootings$manner_of_death)
```

```{r}
#|label: percentage division of the causes of death

table(shootings$manner_of_death) / nrow(shootings) * 100
```

```{r}
#|label: armed status of the victims

table(shootings$armed) / nrow(shootings) * 100
```

```{r}
#|label: sorting the results in decreasing decreasing order

# get percentages of the arms the victims had
temp <- table(shootings$armed) / nrow(shootings) * 100

# sort the result in a decreasing order
temp <- sort(temp, decreasing = TRUE)

# round the results to have 2 decimal points
temp <- round(temp, digits = 2)
```

```{r}
#|label: genders of the victims

table(shootings$gender) / nrow(shootings) * 100
```

```{r}
#|label: race of the victims

temp <- table(shootings$race) / nrow(shootings) * 100
temp <- sort(temp)
temp <- round(temp, digits = 2)
temp
```

## 15.2 Graphing a single numeric variable

```{r}
#|label: load the ggplot2 package

library(ggplot2)
# Warning: package 'ggplot2' was built under R version 4.2.2
```

```{r}
#|label: add data and age variable

ggplot(shootings, aes(x = age))
```

### 15.2.1 Histogram

-   The histogram is a very common type of graph for a single numeric variable. Histograms group a numeric variable into categories and then plot them, with the heights of each bar indicating how common the group is.

```{r}
#|label: histogram to visualize age variable - default bin parameter set to 30

ggplot(shootings, aes(x = age)) +
  geom_histogram()
```

```{r}
#|label: set bin parameter to 15

ggplot(shootings, aes(x = age)) +
  geom_histogram(bins = 15)
```

```{r}
#|label: set bin parameter to 45

ggplot(shootings, aes(x = age)) +
  geom_histogram(bins = 45)
```

```{r}
#|label: display proportions instead of the number

ggplot(shootings, aes(x = age)) +
  geom_histogram(
    aes(y = after_stat(count) / after_stat(sum(count))))
```

### 15.2.2 Density plot

-   Density plots are essentially smoothed versions of histograms.

-   They’re especially useful for numeric variables, which are not integers (integers are whole numbers).

-   They’re also useful when you want to be more precise than a histogram as they are - to simplify - histograms where each bar is very narrow.

-   Note that the y-axis of a density plot is automatically labeled “density” and has very small numbers.

-   Interpreting the y-axis is fairly hard to explain to someone not familiar with statistics so I’d caution against using this graph unless your audience is already familiar with it

```{r}
#|label: density plot

ggplot(shootings, aes(x = age)) +
  geom_density()
```

### 15.2.3 Count graph

-   A count graph is essentially a histogram with a bar for every value in the numeric variable - like a less-smooth density plot.

-   Note that this won’t work well if you have too many unique values so I’d strongly recommend rounding the data to the nearest whole number first if you don’t already have an integer.

-   Our age variable is already rounded so we don’t need to do that. To make a count graph, we add `stat_count()` to the `ggplot()`.

```{r}
#|label: count graph - a single bar for every age in the data

ggplot(shootings, aes(x = age)) +
  stat_count()
```

```{r}
#|label: count graph - proportions

ggplot(shootings, aes(x = age)) +
  stat_count(aes(y = after_stat(count) / after_stat(sum(count))))
```

## 15.3 Graphing a categorical variable

### 15.3.1 Bar graph

```{r}
#|label: bar graph - default is alphabetical order

ggplot(shootings, aes(x = race)) +
  geom_bar()
```

```{r}
#|label: crimes per race - in alphabetical order

table(shootings$race)
```

```{r}
#|label: crimes per race - in ascending order

sort(table(shootings$race))
```

```{r}
#|label: crimes per race - in descending order

sort(table(shootings$race), decreasing = TRUE)
```

```{r}
#|label: label of the races

names(sort(table(shootings$race), decreasing = TRUE))
```

```{r}
#|label: make the race column a factor variable

shootings$race <- factor(shootings$race,
                         levels = names(sort(table(shootings$race), decreasing = TRUE)))
```

```{r}
#|label: rebuilding a bar plot

ggplot(shootings, aes(x = race)) +
  geom_bar()
```

-   Note that all the values that are missing in our data are still reported in the barplot under a column called “NA.”

-   This is not sorted properly since there are more NA values than three of the other values, but NA is still at the far right of the graph.

-   We can change this if we want to make all the NA values an actual character type and call it something like “Unknown.”

```{r}
#|label: flip coordinates - sorted from smallest to largest

ggplot(shootings, aes(x = race)) +
  geom_bar() +
  coord_flip()
```

-   Since it’s flipped, now it’s sorted from smallest to largest.

-   So we’ll need to change the `factor()` code to fix that by making the `decreasing` parameter in `sort()` FALSE.

```{r}
#|label: sorted crimes per races from largest to smallest

shootings$race <- factor(shootings$race,
                         levels = names(sort(table(shootings$race), decreasing = FALSE)))
```

```{r}
#|label: flip coordinates - sorted crimes per races from largest to smallest

ggplot(shootings, aes(x = race)) +
  geom_bar() +
  coord_flip()
```

-   The NA value is now at the top, which looks fairly bad.

-   Let’s change all NA values to the string “Unknown”. And while we’re at it, let’s change all the abbreviated race values to actual names.

```{r}
#|label: convert race variable into character

shootings$race <- as.character(shootings$race)
```

```{r}
#| label: change NAs to unknown

shootings$race[is.na(shootings$race)] <- "Unknown"
```

-   Now we can use conditional statements to change all the race letters to names. I

-   t’s not clear what race “O” and “N” are so I checked [*The Washington Post’s* GitHub](https://github.com/washingtonpost/data-police-shootings) which explains what they mean. Instead of `is.na()` we’ll use `shootings$race == ""`, where we put the letter inside of the quotes.

```{r}
#|label: change race labels to actual race names

shootings$race[shootings$race == "O"] <- "Other"
shootings$race[shootings$race == "N"] <- "Native American"
shootings$race[shootings$race == "A"] <- "Asian"
shootings$race[shootings$race == "H"] <- "Hispanic"
shootings$race[shootings$race == "B"] <- "Black"
shootings$race[shootings$race == "W"] <- "White"
```

```{r}
#|label: convert race variable to factor again

shootings$race <- factor(shootings$race,
                         levels = names(sort((table(shootings$race)), decreasing = FALSE)))
```

```{r}
#|label: create a bar graph again

ggplot(shootings, aes(x = race)) +
  geom_bar() +
  coord_flip()
```

```{r}
#|label: create a bar graph for proportions

ggplot(shootings, aes(x = race)) +
  geom_bar(aes(y = after_stat(count) / after_stat(sum(count)))) +
  coord_flip()
```

## 15.4 Graphing data over time

-   We’ll need to make a variable to indicate that the row is for one shooting.

-   We can call this “dummy” and assign it a value of 1.

-   Then we can make the `ggplot()` and set this “dummy” column to the y-axis value and set our date variable “date” to the x-axis (the time variable is **always** on the x-axis).

-   Then we’ll set the type of plot to `geom_line()` so we have a line graph showing killings over time.

```{r}
#|label: create helper column to indicate that the row is for one shooting

shootings$dummy <- 1
```

```{r}
#|label: create a line graph

ggplot(shootings, aes(x = date, y = dummy)) +
  geom_line()
```

-   This graph is clearly wrong. Why? Well, our y-axis variable is always 1 so there’s no variation to plot. Every single value, even if there are more than one shooting per day, is on the 1 line on the y-axis.

-   And the fact that we have multiple killings per day is an issue because we only want a single line in our graph.

-   We’ll need to aggregate our data to some time period (e.g. day, month, year) so that we have one row per time-period and know how many people were killed in that period.

-   We’ll start with yearly data and then move to monthly data.

```{r}
#|label: load the lubridate package

library(lubridate)
```

-   We’ll use two functions to create variables that tell us the month and the year of each date in our data. We’ll use these new variables to aggregate our data to that time unit.

-   First, the `floor_date()` function is a very useful tool that essentially rounds a date.

-   Here we have the exact date the killing happened on, and we want to determine what month that date is from.

-   So we’ll use the parameter `unit` in `floor_date()` and tell the function we want to know the “month” (for a full set of options please see the documentation for `floor_date()` by entering `?floor_date` in the console).

-   So we can do `floor_date(shootings$date, unit = "month")` to get the month - specifically, it returns the date that is the first of the month for that month - in which the killing happened.

-   Even simpler, to get the year, we simply use `year()` and put our “date” variable in the parentheses. We’ll call the new variables “month_year” and “year”, respectively.

```{r}
#|label: create month and year variables

shootings$month_year <- floor_date(shootings$date, unit = "month")
shootings$year <- year(shootings$date)
```

```{r}
#|label: check the first six values of the month_year variable

head(shootings$month_year)
```

```{r}
#|label: check the first six values of the year variable

head(shootings$year)
```

```{r}
#|label: load the dplyr package

library(dplyr)
```

```{r}
#|label: aggregate the data on month level

monthly_shootings <- shootings |> 
  group_by(month_year) |> 
  summarize(dummy = sum(dummy))
```

```{r}
#|label: check the first six rows of the data set

head(monthly_shootings)
```

```{r}
#|label: a line graph for monthly shootings

ggplot(monthly_shootings, aes(x = month_year, y = dummy)) +
         geom_line()
```

```{r}
#|label: aggregate the data on year level

yearly_shootings <- shootings |> 
  group_by(year) |> 
  summarize(dummy = sum(dummy))
```

```{r}
#|label: a line graph for monthly shootings

ggplot(yearly_shootings, aes(x = year, y = dummy)) +
  geom_line()
```

-   Note the steep drop-off at the end of each graph. Is that due to fewer shooting occurring more recently?

-   No, it’s simply an artifact of the graph comparing whole months (years) to parts of a month (year) since we haven’t finished this month (year) yet.

## 15.5 Pretty graphs

### 15.5.1 Themes

In addition to making changes to the graph’s appearance yourself, you can use a theme that someone else made.

A theme is just a collection of changes to the graph’s appearance that someone put in a function for others to use.

Each theme is different and is fairly opinionated, so you should only use one that you think looks best for your graph.

o use a theme, simply add the function for that theme to your ggplot using the `+` as normal. `ggplot2` comes with a series of themes that you can look at [here](https://ggplot2.tidyverse.org/reference/ggtheme.html).

```{r}
#|label: load the ggtheme package

library(ggthemes)
```

```{r}
#|label: five thirty eight theme

ggplot(yearly_shootings, aes(x = year, y = dummy)) +
  geom_line() +
  theme_fivethirtyeight()
```

```{r}
#|label: tufte theme

ggplot(yearly_shootings, aes(x = year, y = dummy)) +
  geom_line() +
  theme_tufte()
```

```{r}
#|label: few theme

ggplot(yearly_shootings, aes(x = year, y = dummy)) +
  geom_line() +
  theme_few()
```

```{r}
#|label: excel theme

ggplot(yearly_shootings, aes(x = year, y = dummy)) +
  geom_line() +
  theme_excel()
```
