---
title: "Interactive maps"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/08/18
---

-   Data:

    -   san_francisco_marijuana_geocoded.csv

    -   sf_neighborhoods_suicide.rda.

```{r}
#|label: load the readr package

library(readr)
```

```{r}
#|label: read the data into R

marijuana <- read_csv("data/san_francisco_marijuana_geocoded.csv")
marijuana <- as.data.frame(marijuana)
```

## 18.1 Why do interactive graphs matter?

### 18.1.1 Understanding your data

-   The most important thing to learn from this book is that understanding your data is crucial to good research.

-   Making interactive maps is a very useful way to better understand your data as you can immediately see geographic patterns and quickly look at characteristics of those incidents to understand them.

### 18.1.2 Police departments use them

-   Interactive maps are popular in large police departments, such as Philadelphia and New York City.

-   They allow easy understanding of geographic patterns in the data and, importantly, allow such access to people who do not have the technical skills necessary to interact with the data itself.

## 18.2 Making the interactive map

```{r}
#|label: check the first 6 rows of data

head(marijuana)
```

```{r}
#|label: load the leaflet package

library(leaflet)
```

```{r}
#|label: create the first map - world map

leaflet() |> 
  addTiles()
```

```{r}
#|label: display marijuana dispensaries in San Francisco

leaflet() |> 
  addTiles() |> 
  addMarkers(lng = marijuana$long,
             lat = marijuana$lat)
```

```{r}
#|label: change icons to circles

leaflet() |> 
  addTiles() |> 
  addCircleMarkers(lng = marijuana$long,
                   lat = marijuana$lat)
```

```{r}
#|label: change size of circles

leaflet() |> 
  addTiles() |> 
  addCircleMarkers(lng = marijuana$long,
                   lat = marijuana$lat,
                   radius = 5)
```

## 18.3 Adding popup information

```{r}
#|label: adding popup

leaflet() |> 
  addTiles() |> 
  addCircleMarkers(lng = marijuana$long,
                   lat = marijuana$lat,
                   radius = 5,
                   popup = marijuana$Business_Owner)
```

```{r}
#|label: add a descriptive text before business owner

leaflet() |> 
  addTiles() |> 
  addCircleMarkers(lng = marijuana$long,
                   lat = marijuana$lat,
                   radius = 5,
                   popup = paste("Business Owner:", 
                                 marijuana$Business_Owner))
```

```{r}
#|label: add the address and license to the popup

leaflet() |> 
  addTiles() |> 
  addCircleMarkers(lng = marijuana$long,
                   lat = marijuana$lat,
                   radius = 5,
                   popup = paste("Business Owner:",
                                 marijuana$Business_Owner,
                                 "Address:",
                                 marijuana$Premise_Address,
                                 "License:",
                                 marijuana$License_Number))
```

```{r}
#|label: add line breaks to the popup

leaflet() |> 
  addTiles() |> 
  addCircleMarkers(lng = marijuana$long,
                   lat = marijuana$lat,
                   radius = 5,
                   popup = paste("Business Owner:",
                                 marijuana$Business_Owner,
                                 "<br>",
                                 "Address:",
                                 marijuana$Premise_Address,
                                 "<br>",
                                 "License:",
                                 marijuana$License_Number))
```

## 18.4 Dealing with too many markers

-   In our case with only 33 rows of data, turning the markers to circles solves our visibility issue.

-   In cases with many more rows of data, this doesn’t always work.

-   A solution for this is to cluster the data into groups where the dots only show if you zoom in.

```{r}
#|label: cluster the data into groups

leaflet() |> 
  addTiles() |> 
  addCircleMarkers(lng = marijuana$long,
                   lat = marijuana$lat,
                   radius = 5,
                   popup = paste("Business Owner:",
                                 marijuana$Business_Owner,
                                 "<br>",
                                 "Address:",
                                 marijuana$Premise_Address,
                                 "<br>",
                                 "License:",
                                 marijuana$License_Number),
                   clusterOptions = markerClusterOptions())
```

## 18.5 Interactive choropleth maps

-   In Chapter [17](https://crimebythenumbers.com/choropleth-maps.html#choropleth-maps) we worked on choropleth maps which are maps with shaded regions, such as states colored by which political party won them in an election.

-   Here we will make interactive choropleth maps where you can click on a shaded region and see information about that region.

-   We’ll make the same map as before - neighborhoods shaded by the number of suicides.

-   Let’s load the San Francisco suicides-by-neighborhood data that we made earlier.

-   We’ll also want to project it to the standard longitude and latitude projection, otherwise our map won’t work right.

```{r}
#|label: load the simple features library

library(sf)
```

```{r}
#|label: load San Francisco neighborhoods suicide data

load("data/sf_neighborhoods_suicide.rda")
```

```{r}
#|label: change the projection of the map

sf_neighborhoods_suicide <- st_transform(sf_neighborhoods_suicide,
                                         "+proj=longlat +datum=WGS84")
```

```{r}
#|label: create a leaflet map - a map with thick blue lines indicating each neighborhood

# imput is the geometry column
leaflet() |> 
  addTiles() |> 
  addPolygons(data = sf_neighborhoods_suicide$geometry)
```

```{r}
#|label: create a leaflet map - change border colors to black and make them thinner

leaflet() |> 
  addTiles() |> 
  addPolygons(data = sf_neighborhoods_suicide$geometry,
              color = "black",
              weight = 1)
```

```{r}
#|label: create a leaflet map - add popup

leaflet() |>
  addTiles() |> 
  addPolygons(data = sf_neighborhoods_suicide$geometry,
              color = "black",
              weight = 1,
              popup = paste0("Neighborhood: ",
                             sf_neighborhoods_suicide$nhood,
                             "<br>",
                             "Number of Suicides: ",
                             sf_neighborhoods_suicide$number_suicides))
```

-   For these types of maps we generally want to shade each polygon to indicate how frequently the event occurred in the polygon.

-   We’ll use the function `colorNumeric()`, which takes a lot of the work out of the process of coloring in the map.

-   This function takes two inputs, first a color palette, which we can get from the site [Color Brewer.](http://colorbrewer2.org/#type=sequential&scheme=OrRd&n=3)

-   The second parameter is the column for our numeric variable, *number_suicides*.

-   We will save the output of `colorNumeric("OrRd", sf_neighborhoods_suicide$number_suicides)` as a new object, which we’ll call *pal* for convenience since it is a palette of colors.

-   Then inside of `addPolygons()` we’ll set the parameter `fillColor` to `pal(sf_neighborhoods_suicide$number_suicides)`, running this function on the column.

-   What this really does is determine which color every neighborhood should be based on the value in the *number_suicides* column.

```{r}
#|label: create a color palette object

pal <- colorNumeric("OrRd", sf_neighborhoods_suicide$number_suicides)
```

```{r}
#|label: create a leaflet map - add a color palette

leaflet() |> 
  addTiles() |> 
  addPolygons(data = sf_neighborhoods_suicide$geometry,
              color = "black",
              weight = 1,
              popup = paste0("Neighborhood: ",
                             sf_neighborhoods_suicide$nhood,
                             "<br>",
                             "Number of Suicides: ",
                             sf_neighborhoods_suicide$number_suicides),
              fillColor = pal(sf_neighborhoods_suicide$number_suicides))
```

```{r}
#|label: create a leaflet map - make each neighborhood a solid color

leaflet() |> 
  addTiles() |> 
  addPolygons(data = sf_neighborhoods_suicide$geometry,
              color = "black",
              weight = 1,
              popup = paste0("Neighborhood: ",
                             sf_neighborhoods_suicide$nhood,
                             "<br>",
                             "Number of Suicides: ",
                             sf_neighborhoods_suicide$number_suicides),
              fillColor = pal(sf_neighborhoods_suicide$number_suicides),
              fillOpacity = 1)
```

```{r}
#|label: create a leaflet map - add a legend

leaflet() |> 
  addTiles() |> 
  addPolygons(data = sf_neighborhoods_suicide$geometry,
              color = "black",
              weight = 1,
              popup = paste0("Neighborhood: ",
                             sf_neighborhoods_suicide$nhood,
                             "<br>",
                             "Number of Suicides: ",
                             sf_neighborhoods_suicide$number_suicides),
              fillColor = pal(sf_neighborhoods_suicide$number_suicides),
              fillOpacity = 1) |> 
  addLegend(pal = pal,
            values = sf_neighborhoods_suicide$number_suicides,
            opacity = 1)
```

```{r}
#|label: create a leaflet map - add a legend title

leaflet() |> 
  addTiles() |> 
  addPolygons(data = sf_neighborhoods_suicide$geometry,
              color = "black",
              weight = 1,
              popup = paste0("Neighborhood: ",
                             sf_neighborhoods_suicide$nhood,
                             "<br>",
                             "Number of Suicides: ",
                             sf_neighborhoods_suicide$number_suicides),
              fillColor = pal(sf_neighborhoods_suicide$number_suicides),
              fillOpacity = 1) |> 
  addLegend(pal = pal,
            values = sf_neighborhoods_suicide$number_suicides,
            opacity = 1,
            title = "Suicides" ) |> 
  addProviderTiles(providers$CartoDB.Positron)
```
