---
title: "More scraping tables from PDFs"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/08/29
---

Data:

-   AbbreRptCurrent.pdf

-   PregnantFemaleReportingCurrent.pdf.

## 23.1 Texas jail data

### Package Loading

**Purpose**: Comments showing which packages to install if not already available.

-   `pdftools`: Extracts text from PDF files

-   `stringr`: String manipulation functions

-   `dplyr`: Data manipulation and transformation

-   `readr`: Reading and writing data files

```{r}
#|label: load the necessary packages for the Texas jail data

library(pdftools)
library(stringr)
library(dplyr)
library(readr)
```

### File Path Setup

**Purpose**: Store the path to your PDF file in a variable for easy reference throughout the script.

```{r}
#|label: set your PDF file path

pdf_file_path <- "data/AbbreRptCurrent.pdf"
```

### File Existence Check

**Purpose**: Verify the PDF file exists before trying to process it.

-   `file.exists()`: Returns TRUE if file exists, FALSE if not

-   `!`: Negation operator (NOT)

-   `getwd()`: Gets current working directory

-   `stop()`: Halts execution with error message if file not found

```{r}
#|label: check if file exists

if (!file.exists(pdf_file_path)) {
  cat("✗ File not found:", pdf_file_path, "\n")
  cat("Current directory:", getwd(), "\n")
  stop("Please check your file path")
} else {
  cat("✓ File found:", pdf_file_path, "\n")
}
```

### PDF Information Extraction

**Purpose**: Extract basic metadata about the PDF.

-   `pdf_info()`: Returns information about the PDF (pages, title, creator, etc.)

-   `$`: Accesses specific elements from the info object

```{r}
#|label: get PDF info

pdf_info <- pdf_info(pdf_file_path)
cat("PDF Pages:", pdf_info$pages, "\n")
cat("PDF Title:", pdf_info$title, "\n")
cat("PDF Creator:", pdf_info$creator, "\n")
```

### Text Extraction with Error Handling

**Purpose**: Attempt to extract all text from the PDF with error handling.

-   `tryCatch()`: Executes code and catches any errors

-   `pdf_text()`: Extracts text from all pages of PDF, returns vector of strings (one per page)

-   `error = function(e)`: What to do if an error occurs

-   Returns `NULL` if extraction fails

```{r}
#|label: try to extract text using pdftools

text_content <- tryCatch({
  pdf_text(pdf_file_path)
}, error = function(e) {
  cat("Error extracting text:", e$message, "\n")
  NULL
})
```

### Text Content Analysis

**Purpose**: Check if text extraction was successful.

-   `is.null()`: Checks if variable is NULL

-   `length()`: Number of elements in the vector (= number of pages)

**Purpose**: Count total characters to assess if meaningful text was found.

-   `nchar()`: Number of characters in each string

-   `sum()`: Add up character counts from all pages

**Purpose**: Determine if PDF contains enough text to be considered text-based (not scanned image).

### Text Preview

**Purpose**: Show preview of extracted text from first page.

-   `text_content[1]`: First element (first page) of the vector

-   `substr()`: Extract substring from position 1 to 1000

### Text Processing

**Purpose**: Combine all pages into single string.

-   `paste()`: Combine strings

-   `collapse = "\n"`: Join with newline characters between pages

**Purpose**: Split the combined text into individual lines.

-   `strsplit()`: Split string at specified pattern ("\\n" = newlines)

-   `[[1]]`: Extract first (and only) element from the result list

**Purpose**: Remove empty lines.

-   `trimws()`: Remove whitespace from beginning and end of strings

-   `nchar() > 0`: Keep only lines with at least one character

-   `lines[...]`: Subset lines based on the condition

### Data Line Identification

**Purpose**: Find lines that look like data (contain multiple numbers separated by spaces).

-   `grepl()`: Returns TRUE/FALSE for each line matching the pattern

-   `"\\d+\\s+\\d+"`: Regular expression pattern:

    -   `\\d+`: One or more digit

    -   `\\s+`: One or more white space characters

    -   Pattern finds lines with at least two numbers separated by spaces

### Data Parsing Function

**Purpose**: Function to parse a single data line into columns.

-   `trimws()`: Remove leading/trailing white space

-   `strsplit(line, "\\s{2,}")`: Split on 2 or more consecutive spaces

-   `unlist()`: Convert list result to vector

-   Remove empty parts and return

### Apply Parsing to All Data Lines

**Purpose**: Apply the parsing function to every data line.

-   `lapply()`: Apply function to each element of a list/vector

-   Returns list where each element is a parsed line

### Column Consistency Check

**Purpose**: Find the most common number of columns to handle inconsistent parsing.

-   `sapply()`: Apply `length()` to each parsed line

-   `table()`: Count how many times each column count appears

-   `sort(..., decreasing = TRUE)`: Sort counts from highest to lowest

-   `[1]`: Take the most common count

-   `names()`: Get the column count value

-   `as.numeric()`: Convert to number

### Data Frame Creation

**Purpose**: Keep only rows with the most common number of columns.

**Purpose**: Convert list of parsed lines into a matrix.

-   `lapply(..., function(x))`: Apply function to each row

-   `if (length(x) == most_common_cols)`: Check if row has right number of columns

-   `rep(NA, most_common_cols)`: Create row of missing values if wrong length

-   `do.call(rbind, ...)`: Combine all rows into matrix using `rbind()`

**Purpose**: Convert matrix to data frame, keeping text as character strings (not factors).

### Column Naming

**Purpose**: Assign meaningful column names if we have the expected number of columns.

-   `c()`: Create vector of column names

-   `ncol()`: Number of columns in data frame

-   `names(texas_jail) <-`: Assign column names

-   `paste0("V", 1:ncol(texas_jail))`: Create generic names V1, V2, V3, etc.

### Data Type Conversion

**Purpose**: Convert numeric columns from text to numbers.

-   `names(texas_jail)[-1]`: All column names except the first (County)

-   `for (col in numeric_cols)`: Loop through each numeric column

-   `gsub("[^0-9.-]", "", ...)`: Remove all non-numeric characters except digits, periods, and hyphens

-   `as.numeric()`: Convert cleaned text to numbers

-   `texas_jail[[col]] <-`: Assign cleaned values back to column

### Data Cleaning

**Purpose**: Remove header rows that got mixed in with data.

-   `grepl(..., ignore.case = TRUE)`: Find rows where first column contains "County", "Total", or "Page"

-   `!`: Negate (keep rows that DON'T match)

-   `texas_jail[..., ]`: Subset rows based on condition

**Purpose**: Remove rows with empty county names.

-   `is.na()`: Check for missing values

-   `texas_jail[,1] != ""`: Check for empty strings

-   `&`: AND operator (both conditions must be true)

-   `!`: Keep rows where county name is NOT missing AND NOT empty

### Save Results

**Purpose**: Save the cleaned data in two formats.

-   `write_csv()`: Save as CSV file (readable by Excel, other programs)

-   `saveRDS()`: Save as R data file (preserves data types, faster loading in R)

```{r}
#|label: get final clean data frame

if (!is.null(text_content)) {
  cat("✓ Text extraction successful\n")
  cat("Number of pages with text:", length(text_content), "\n")
  
  
############
  total_chars <- sum(nchar(text_content))
  cat("Total characters extracted:", total_chars, "\n")
  
  
############
# Purpose: Determine if PDF contains enough text to be considered text-based (not scanned image).
  if (total_chars > 100) {
    cat("✓ Substantial text content found - PDF is text-based\n")
    

############
    cat("\n=== FIRST PAGE TEXT PREVIEW ===\n")
    first_page <- text_content[1]
    cat(substr(first_page, 1, 1000), "\n")
    cat("...[truncated]\n")

    
############
    all_text <- paste(text_content, collapse = "\n")
    
  
############
    lines <- strsplit(all_text, "\n")[[1]]


############
    # Remove empty lines
    lines <- lines[nchar(trimws(lines)) > 0]
    cat("Total non-empty lines:", length(lines), "\n")

    
############
    # Look for lines that might be data rows (contain county names and numbers)
    # Based on your document, look for lines with multiple numbers
    data_lines <- lines[grepl("\\d+\\s+\\d+", lines)]
    
    cat("Lines that look like data rows:", length(data_lines), "\n")
    
    if (length(data_lines) > 0) {
      cat("\nFirst few data lines:\n")
      print(head(data_lines, 10))


############
      # Attempt to parse the structured data
      cat("\n=== PARSING STRUCTURED DATA ===\n")


############
      # Function to parse a data line
      parse_jail_line <- function(line) {
        # Clean the line
        line <- trimws(line)
        
        # Split by multiple spaces (assuming columns are separated by multiple spaces)
        parts <- unlist(strsplit(line, "\\s{2,}"))
        
        # Remove empty parts
        parts <- parts[nchar(trimws(parts)) > 0]
        
        return(parts)
      }
      
      
############
      # Parse all data lines
      parsed_data <- lapply(data_lines, parse_jail_line)
      

############
      # Find the most common number of columns (to handle inconsistencies)
      col_counts <- sapply(parsed_data, length)
      most_common_cols <- as.numeric(names(sort(table(col_counts), decreasing = TRUE)[1]))
      cat("Most common number of columns:", most_common_cols, "\n")


#############
      # Filter to rows with the most common column count
      consistent_rows <- parsed_data[col_counts == most_common_cols]
      

#############
      if (length(consistent_rows) > 0) {
        # Convert to data frame
        texas_jail <- do.call(rbind, lapply(consistent_rows, function(x) {
          if (length(x) == most_common_cols) {
            return(x)
          } else {
            return(rep(NA, most_common_cols))
          }
        }))
        

#############
        texas_jail <- as.data.frame(texas_jail, stringsAsFactors = FALSE)
   
         
#############
        # Set column names (you may need to adjust these based on your data)
        expected_columns <- c(
          "county",
          "pretrial_felons",
          "conv_felons", 
          "conv_felons_sentenced_to_county_jail_time",
          "parole_violators",
          "parole_violators_with_a_New_Charge",
          "pretrial_misd",
          "conv_misd",
          "bench_warrants",
          "federal",
          "pretrial_sjf",
          "conv_sjf_sentenced_to_co_jail_time",
          "conv_sjf_sentenced_to_state_jail",
          "total_others",
          "total_local", 
          "total_contract",
          "total_population",
          "total_capacity",
          "percent_of_capacity",
          "available_beds"
        )
        
        if (ncol(texas_jail) == length(expected_columns)) {
          names(texas_jail) <- expected_columns
        } else {
          names(texas_jail) <- paste0("V", 1:ncol(texas_jail))
        }
        
        cat("Parsed data frame dimensions:", nrow(texas_jail), "x", ncol(texas_jail), "\n")
        

#############
        # Clean numeric columns
        numeric_cols <- names(texas_jail)[-1]
        for (col in numeric_cols) {
          if (col %in% names(texas_jail)) {
            texas_jail[[col]] <- as.numeric(gsub("[^0-9.-]", "", texas_jail[[col]]))
          }
        }
        
#############
        # Remove rows where county name looks like headers
        texas_jail <- texas_jail[!grepl("County|Total|Page", 
                                                  texas_jail[,1], 
                                                  ignore.case = TRUE), ]
        

#############
        # Remove empty rows
        texas_jail <- texas_jail[!is.na(texas_jail[,1]) & 
                                             texas_jail[,1] != "", ]
        
        cat("Final cleaned data:", nrow(texas_jail), "rows\n")
        

#############
        if (nrow(texas_jail) > 0) {
          # Save results
          write_csv(texas_jail, "texas_jail_population_parsed.csv")
          saveRDS(texas_jail, "texas_jail_population_parsed.rds")
          
          cat("✓ Data saved to:\n")
          cat("- texas_jail_population_parsed.csv\n")
          cat("- texas_jail_population_parsed.rds\n")
          
          # Show preview
          cat("\nFirst 6 rows of parsed data:\n")
          print(head(texas_jail))
          
          cat("\n✓ Data available in 'jail_data' variable\n")
          
        } else {
          cat("✗ No valid data rows after cleaning\n")
        }
      } else {
        cat("✗ No consistent data rows found\n")
      }
    } else {
      cat("✗ No data lines detected in the text\n")
    }
  }
} else {
  cat("✗ Could not extract any text from PDF\n")
  cat("This confirms the PDF is likely image-based (scanned)\n")
}
```

## 23.2 Pregnant women incarcerated

### Load libraries

```{r}
#|label: load the necessary packages Pregnant Female Booked Data

library(stringr)
library(tidyr)
library(dplyr)
library(pdftools)
```

### File Path Setup

```{r}
#|label: file path set up

pdf_file_path <- "data/PregnantFemaleReportingCurrent.pdf"
```

### File Existence Check

```{r}
#|label: check if the file exists

if(!file.exists(pdf_file_path)) {
  cat("✗ File not found:", pdf_file_path, "\n")
  cat("Current directory:", getwd(), "\n")
  stop("Please check your file path")
} else {
  cat("✓ File found:", pdf_file_path, "\n")
}
```

```{r}
#|label: get file information

pdf_info <- pdf_info(pdf_file_path)
cat("PDF Pages:", pdf_info$pages, "\n")
```

### Text Extraction with Error Handling

```{r}
#|label: attempt to extract all text from the PDF with error handling

text_content <- tryCatch({
  pdf_text(pdf_file_path)
}, error = function(e) {
  cat("Error extracting text:", e$message)
})
```

### Text Content Analysis

```{r}
#|label: check if text extraction was successful

if(!is.null(text_content)) {
  cat("✓ Text extraction successful\n")
  cat("Number of pages with text:", length(text_content), "\n")
}
```

```{r}
#|label: count total characters to assess if meaningful text was found

total_chars <- sum(nchar(text_content))
cat("Total characters extracted: ", total_chars)
```

```{r}
#|label: determine if PDF contains enough text to be considered text-based (not scanned image)

if(total_chars > 100) {
  cat("✓ Substantial text content found - PDF is text-based\n")
}
```

### Text Preview

```{r}
#|label: check the raw extracted text

text_content
```

### Data Cleaning

```{r}
#|label: remove the first line

text_content <- str_replace(text_content,
                            "Pregnant Females Booked.*?Month Total:\\s*\\d+\\n",
                            "")
```

```{r}
#|label: check data after removing the title

text_content
```

```{r}
#|label: split into lines

lines <- strsplit(text_content, "\n")[[1]]
```

```{r}
#|label: split into columns

lines <- lines[str_trim(lines) != ""]
```

```{r}
#|label: check data again after splitting into lines and columns

lines
```

```{r}
#|label: create a matrix from split data

final_data_matrix <- do.call(rbind, lapply(lines, function(line) {
  cols <- str_split(str_trim(line), "\\s{2,}")[[1]]
  length(cols) <- 10 # Ensure 10 colums
  return(cols)
}))
```

```{r}
#|label: convert to data frame

final_data_converted <- as.data.frame(final_data_matrix, stringsAsFactors = FALSE)
```

```{r}
#|label: check data frame after conversion

head(final_data_converted)
```

```{r}
#|label: reshape to long format

pregnant_female_booked <- data.frame()

for (i in seq(1, 9, 2)) {
  temp <- data.frame(
    county = final_data_converted[[i]],
    pregnant_females_booked = as.numeric(final_data_converted[[i+1]])
  )
  temp <- temp[!is.na(temp$county) 
               & temp$county != "" 
               & !is.na(temp$pregnant_females_booked), ]
  pregnant_female_booked <- rbind(pregnant_female_booked, temp)
}
```
