---
title: "R Markdown"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/07/2025
---

-   File \> New File \> R Markdown

-   Title \| Author \| Default Output Format \> Ok

## 7.1 Code

Insert (or CTRL (Command) + ALT + I) \> R

### 7.1.1 Hiding code in the output

```{r echo=FALSE}
summary(cars)
```

## 7.2 Inline Code

The data set mtcars has `r nrow(mtcars)` rows and `r ncol(mtcars)` columns. The mean of the mpg column is `r mean(mtcars$mpg)`.

## 7.3 Tables

```{r}
#| label: load the knitr package

library(knitr)
```

```{r}
#|label: subset the data set

mtcars_small <- mtcars[1:5, 1:5]
mtcars_small
```

```{r}
#|label: make a table

kable(mtcars_small, caption = "This is an example caption")
```

## 7.4 Footnotes

-   Insert \> Footnote

-   Explain this sentence in a footnote[^1]

[^1]: This is an example of a footnote.

## 7.5 Citation

-   Insert \> Citation

## 7.6 Spell check

-   ABC

## 7.7 Making the output file

-   Knit \>

    -   Knit to HTML

    -   Knit to PDF

    -   Knit to Word
