---
title: "Git"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/07/18
---

# 9.1 What is Git, and why do I need it?

-   Git is the way R and many other programming languages (technically any type of files and folders) does the "version control"

# 9.2 Git basics

-   Four main processes:

    -   checkout

    -   add and commit

    -   push

    -   pull

+--------------------------------------------------------------------+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Library Steps**                                                  | **Git steps**                                       | **Git code**                                                                                                                                                                      |
+====================================================================+=====================================================+===================================================================================================================================================================================+
| Go to library                                                      |                                                     |                                                                                                                                                                                   |
+--------------------------------------------------------------------+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Find book and check out book                                       | Clone (usually will just be done once per project). | Git clone *path to repo, can be GitHub link*                                                                                                                                      |
+--------------------------------------------------------------------+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Read or write in book                                              | This is done in R, not in Git                       | No Git code, this is going to be whatever code we write in R. Also includes any outputs such as making a graph that is saved, R Markdown outputs like a PDF, or even new R files. |
+--------------------------------------------------------------------+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Return book                                                        | Add and commit\                                     | Git add .\                                                                                                                                                                        |
|                                                                    | Push                                                | Git commit –m “message indicating what we wrote”\                                                                                                                                 |
|                                                                    |                                                     | Git push                                                                                                                                                                          |
+--------------------------------------------------------------------+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Check out book again (to see what other people have written in it) | Pull                                                | Git pull                                                                                                                                                                          |
+--------------------------------------------------------------------+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

# 9.3 Using Git

### 9.3.1 Setting up Git

```{r}
#|label: load the usethis package

library(usethis)
```

```{r}
#|label: connect to GitHub

use_git_config(user.name = "mardan-mirzaguliyev",
               user.email = "mardan.mirzaguliyev@gmail.com")
```

### 9.3.2 Setting up GitHub

-   <https://github.com/mardan-mirzaguliyev/cbtn>

-   Watch/Unwatch

-   Star

-   Fork

# 9.4 Setting up Git on an already-made R Project

-   Tools \> Project Options \> Git/SVN \> Git \> Yes \> Restart \> Yes

# 9.5 Using Git through RStudio
