---
title: "19 Webscraping with rvest"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/08/23
---

```{r}
#|label: load the rvest package

library(rvest)
```

-   [**MMMMM... Brownies**](https://www.allrecipes.com/recipe/25080/mmmmm-brownies/)

## 19.1 Scraping one page

```{r}
#|label: read html structure of the web page

brownies <- read_html("https://www.allrecipes.com/recipe/25080/mmmmm-brownies/")
```

-   We need to find just which parts of the page to scrape.

-   To do so we’ll use the helper tool [SelectorGadget](https://selectorgadget.com/), a Google Chrome extension that lets you click on parts of the page to get the CSS selector code that we’ll use.

-   Install that extension in Chrome and go to the [brownie recipe page.](https://www.allrecipes.com/recipe/25080/mmmmm-brownies/?internalSource=hub%20recipe&referringContentType=Search)

### STEP 1. Get html nodes using CSS selector

```{r}
#|label: get ingredients of the pastry

ingredients <- html_nodes(brownies, 'span[data-ingredient-name="true"]')
```

### STEP 2. Load the data as a text

```{r}
#|label: express the format of the scraped data for ingredients - text

ingredients <- html_text(ingredients)
```

```{r}
#|label: check ingredients data

ingredients
```

### STEP 3. Use the same process to get the directions for baking - Get html nodes using CSS selector

```{r}
#|label: get directions for baking

directions <- html_nodes(brownies, '.mntl-sc-block-group--LI .mntl-sc-block')
```

### STEP 4. Load the data as a text

```{r}
#|label: express the format of the scraped data for baking directions - text

directions <- html_text(directions)
```

```{r}
#|label: check directions data

directions
```

## 19.2 Cleaning the webscraped data

### STEP 5. Clean the webscraped data

```{r}
#|label: remove spaces

ingredients <- trimws(ingredients)
directions <- trimws(directions)
```

```{r}
#|label: check ingredients data for spaces

ingredients
```

```{r}
#|label: check directions data for spaces

directions
```

```{r}
#|label: remove spaces between the step number and the instructions

directions <- gsub(" {2,}", "", directions)
```

```{r}
#|label: final check for directions

directions
```

```{r}
#|label: AI generated code for correct webscraping of ingredients data

library(rvest)
library(jsonlite)

# Get the page
page <- read_html("https://www.allrecipes.com/recipe/25080/mmmmm-brownies/")

# Look for JSON-LD structured data
json_scripts <- html_nodes(page, 'script[type="application/ld+json"]')

# Extract recipe data
if(length(json_scripts) > 0) {
  recipe_data <- fromJSON(html_text(json_scripts[[1]]))
  
  # Get full ingredients with quantities
  full_ingredients <- recipe_data$recipeIngredient
  print(full_ingredients)
  
  # Get directions too
  directions <- recipe_data$recipeInstructions$text
  print(directions)
}
```

```{r}
# Clean your existing directions to remove image captions
clean_directions <- function(directions) {
  # Remove short text (likely captions)
  directions <- directions[nchar(directions) > 20]
  
  # Remove obvious image captions
  directions <- directions[!grepl("^(Photo|Image|Step \\d+ photo)", directions, ignore.case = TRUE)]
  
  # Keep only text with cooking words
  cooking_keywords <- c("mix", "add", "stir", "bake", "cook", "heat", "pour", "combine", "whisk", "fold", "preheat", "grease", "beat", "blend")
  
  valid_directions <- c()
  for(direction in directions) {
    has_cooking_word <- any(sapply(cooking_keywords, function(keyword) {
      grepl(keyword, direction, ignore.case = TRUE)
    }))
    
    if(has_cooking_word || nchar(direction) > 50) {
      valid_directions <- c(valid_directions, direction)
    }
  }
  
  # Add step numbers
  if(length(valid_directions) > 0) {
    valid_directions <- paste0("Step ", 1:length(valid_directions), ": ", valid_directions)
  }
  
  return(valid_directions)
}

# Apply to your scraped directions
cleaned_directions <- clean_directions(directions)
print(cleaned_directions)
```
