---
title: "Data types and structures"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025-06-27
---

## 3.1 Data types

```{r}
#|label: number as a text

# a <- "2"
# b <- "2"
# 
# a + b
# 
# Error in a + b : non-numeric argument to binary operator
```

```{r}
#|label: correct number format

a <- 2
b <- 2

a + b
```

## 3.3 Numeric, character, and logical (boolean)

```{r}
#|label: Usage of boolean type

2 == 3
```

-   **`is()` function**

```{r}
#|label: example 1

is(2)
```

```{r}
#|label: example 2

is("2")
```

```{r}
#|label: example 3

is(TRUE)
```

```{r}
#|label: example 4

is("TRUE")
```

```{r}
#|label: example 5

# is(true)

# Error: object 'true' not found
```

-   **`class()` function**

```{r}
#|label: class() function examples

class(2)
class("2")
class(TRUE)
```

-   **`is.numeric()`**

-   **`is.character()`**

-   **`is.logical()`** functions

```{r}
#|label: examples

is.numeric(2)
is.character("2")
is.logical(TRUE)
```

-   [**The Uniform Crime Reporting (UCR)**](https://en.wikipedia.org/wiki/Uniform_Crime_Reports) program compiles official data on crime in the United States, published by the Federal Bureau of Investigation (FBI). UCR is "a nationwide, cooperative statistical effort of nearly 18,000 city, university and college, county, state, tribal, and federal law enforcement agencies voluntarily reporting data on crimes brought to their attention".

```{r}
#|label: load the data

load("data/ucr2017.rda")
```

-   **`names()`**

-   **`colnames()`** functions

```{r}
#|label: get column names

names(ucr2017)
```

```{r}
#|label: check the type of year column 1

is(ucr2017$year)
```

```{r}
#|label: check the type of year column 2

is.numeric(ucr2017$year)
```

## 3.3 Data structures

### 3.3.1 Vectors (collections of "things")

```{r}
#|label: creating a vector

a <- c(1, 2, 3)
a
```

```{r}
#|label: Errors without c() function and comma

# a <- 1, 2, 3
# Error: unexpected ',' in "a <- 1,"

# a <- c(1 2 3)
# Error: unexpected numeric constant in "a <- c(1 2"
```

```{r}
#|label: vectors must have elements of the same type

c("cat", "dog", "2")
```

### 3.3.2 Data frames

```{r}
#|label: creating a data frame

example <- data.frame(colum1 = c(1, 3, 5, 7, 9),
                      column2 = c("hello",
                                  "darkness",
                                  "my",
                                  "old",
                                  "friend"))

example
```

```{r}
#|label: creating a data frame without naming columns

example <- data.frame(c(1, 3, 5, 7, 9),
                      c("hello",
                        "darkness",
                        "my",
                        "old",
                        "friend"))

example
```

```{r}
#|label: creating a data frame with vectors already made: using default column name and overwriting column name

column_1 <- c(1, 3, 5, 7, 9)
column_2 <- c("hello", "darkness", "my", "old", "friend")

example <- data.frame(column_1, overridden_name = column_2)

example
```

```{r}
#|label: applying is() function

is(example)
```

-   **`nrow()`**,**`ncol()`**, **`dim()`** functions

```{r}
#|label: applying nrow() and ncol() function

nrow(example)

ncol(example)
```

```{r}
#|label: applying dim() function: the first value returned is the number of rows, the second value is the number of columns

dim(example)
```

### 3.3.3 Other data structures

-   **List**

    -   A list is essentially a vector but where different values can be different types

    -   All data frames are actually lists. They are a list of vectors.

```{r}
#|label: creating a list

list_example <- list(
                    # number
                    0,
                    
                    # character
                    "hello",
                    
                    # logical
                    TRUE,
                    
                    # numeric vector
                     1:5,
                    
                    # matrix
                    matrix(1:50, nrow = 10, ncol = 5),
                    
                    # list of numeric vector
                    list(c(33, 66, 99)),
                    
                    # data frame
                    head(mtcars))
  
head(list_example)
```

-   **Matrix**

    -   A matrix is a two-dimensional object where every value is the same type.

    -   Think of a data.frame but each column has to be the same type.

```{r}
#|label: creating a matrix

matrix(1:50, nrow = 5, ncol = 5)
```

```{r}
#|label: matrix should have the same type of values

matrix(c("1", 2:250), nrow = 5, ncol = 5)
```

```{r}
matrix(1:50, nrow = 10, ncol = 5)
```
