---
title: "19 Webscraping with rvest"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/08/23
---

```{r}
#|label: load the rvest and jsonlite packages

library(rvest)
```

-   [**MMMMM... Brownies**](https://www.allrecipes.com/recipe/25080/mmmmm-brownies/)

## 19.1 and 19.2 Scraping and cleaning the webscarepd data one page

### ALL STEPS MERGED BECAUSE OF THE CHANGE OF THE HTML STRUCTURE

```{r}
#|label: read html structure of the web page

brownies <- read_html("https://www.allrecipes.com/recipe/25080/mmmmm-brownies/")
```

-   We need to find just which parts of the page to scrape.

-   To do so we’ll use the helper tool [SelectorGadget](https://selectorgadget.com/), a Google Chrome extension that lets you click on parts of the page to get the CSS selector code that we’ll use.

-   Install that extension in Chrome and go to the [brownie recipe page.](https://www.allrecipes.com/recipe/25080/mmmmm-brownies/?internalSource=hub%20recipe&referringContentType=Search)

#### 1. Get and clean ingredients data

```{r}
#|label: get html nodes using CSS selector

ingredients <- html_nodes(brownies, ".mm-recipes-structured-ingredients__list")
```

```{r}
#|label: load the data as a text

ingredients <- html_text(ingredients)
```

```{r}
#|label: check ingredients data

ingredients
```

```{r}
#|label: remove extra spaces

ingredients <- trimws(ingredients)
ingredients
```

```{r}
#|label: remove \n characters from the vector

ingredients <- gsub("\n\n\n", " ", ingredients)
ingredients
```

```{r}
#|label: split at measurement patterns

ingredients <- gsub("\\s+(\\d+\\s+tablespoons?)\\s+", "\n\\1 ", ingredients)
ingredients <- gsub("\\s+(\\d+\\s+½\\s+cups?)\\s+", "\n\\1 ", ingredients)
ingredients <- gsub("\\s+(\\d+\\s+large)\\s+", "\n\\1 ", ingredients)
ingredients <- gsub("\\s+(½\\s+teaspoon)\\s+", "\n\\1 ", ingredients)
ingredients <- gsub("\\s+(⅔\\s+cup)\\s+", "\n\\1 ", ingredients)
ingredients <- gsub("\\s+(¼\\s+teaspoon)\\s+", "\n\\1 ", ingredients)
```

```{r}
#|label: split vector elements using newline

ingredients <- strsplit(ingredients, "\n")[[1]]
ingredients
```

```{r}
#|label: display ingredients in better format

for(i in 1:length(ingredients)) {
  cat(i, ".", ingredients[i], "\n")
}
```

#### 2. Get and clean directions data

```{r}
#|label: get directions for baking

directions <- html_nodes(brownies, '#mntl-sc-block_1-0')
```

```{r}
#|label: express the format of the scraped data for baking directions - text

directions <- html_text(directions)
```

```{r}
#|label: check directions data

directions
```

```{r}
#|label: remove image caption

directions <- gsub("Dotdash Meredith Food Studios", "", directions, ignore.case = TRUE)
directions
```

```{r}
#|label: remove excessive newlines and spaces

directions <- gsub("\\n+", " ", directions)  # Replace newlines with spaces
directions <- gsub("\\s+", " ", directions)  # Replace multiple spaces with single
directions <- trimws(directions)             # Remove leading/trailing spaces

directions
```

```{r}
#|label: split only at major cooking action words

major_cooking_actions <- c("Gather", "Preheat", "Combine sugar", "Bake", "Let cool")
action_pattern <- paste0("(?<=\\.)\\s+(?=", paste(major_cooking_actions, collapse = "|"), ")")
steps <- strsplit(directions, action_pattern, perl = TRUE)[[1]]

steps
```

```{r}
#|label: clean and filter steps

steps <- trimws(steps)
steps <- steps[nchar(steps) > 15]  # Keep only meaningful steps
steps
```

```{r}
#|label: add the word Step and step numbers

directions <- paste0("Step ", 1:length(steps), ": ", steps)
directions
```

```{r}

#|label: display results in a clean, tidy format

for(i in 1:length(directions)) {
  cat(directions[i], "\n\n")
}
```
