---
title: "Introduction to R and RStudio"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/06/21
---

# 2.1 Using RStudio

Writing a code in an R Scripts keep the code written in a file and this file can be:

1.  Opened later

2.  Edited

3.  Shared with others

## 2.1.1 Opening an R Script

-   File \> New File \> R Script

-   File \> New File \> R Presentation

-   File \> New File \> R Markdown

-   File \> New File \> Quarto Document

## 2.1.2 Setting the working directory

-   Session \> Set Working Directory \>

    -   To Project Directory

    -   To Source File Location

    -   To Files Pane Location

    -   Choose Directory

-   `setwd("path")`

## 2.1.3 Changing RStudio

-   Tools \> Global Options to open up a window with a number of different tabs to change how R behaves and how it looks.

### 2.1.3.1 General

-   Tools \> Global Options \> General \> Uncheck the “Restore .RData into workspace at startup” section \> set “Save workspace to .RData on exit:” section to "Never".

### 2.1.3.2 Code

-   Tools \> Global Options \> Code \> Editing \> Check "Soft-wrap R source files" section

#### 2.1.3.2.1 Saving

-   Tools \> Global Options \> Code \> Saving \> Check “Automatically save when editor loses focus” section and Set time

### 2.1.3.3 Appearance

-   Tools \> Global Options \> Appearance

### 2.1.3.4 Pane Layout

-   Tools \> Global Options \> Page Layout

## 2.1.4 Helpful Cheatsheets

-   Help \> Cheet Sheets

# 2.2 Assigning variables

```{r}
#|label:  Assiging variable

a <- 1
a
```

```{r}
#|:label: Overwriting the value of a variable

a <- 33
a
```

```{r}
#|label: Using `=` sign instead of `<=`

a = 13
a
```

```{r}
#|label: Example demo: Assigning variable to another variable

example_123_value.demonstration <- a
example_123_value.demonstration
```

```{r}
#|label: Changing the value of the variable

example_123_value.demonstration <- example_123_value.demonstration * 10
example_123_value.demonstration
```

# 2.3 What are functions (and packages)?

```{r}

```

-   A function is a bunch of code (it could range from a single line of code to hundreds of lines) that has been assigned to an object.

-   A package is just the name for a collection of functions in an easily downloadable format. 

```{r}
#|label: loading a library

library(caesar)
```

```{r}
#|label: examplple text

caesar("example text")
```

# 2.4 Reading data into R

## 2.3.1 Loading data

```{r}
#|label: loading data

load("data/ucr2017.rda")
```

# 2.5 First steps to exploring data

-   Four functions to explore data:

    -   `head()`

    -   `summary()`

    -   `plot()`

    -   `View()`

## `head()` funcition

```{r}
#|:label: head function example

head(ucr2017)
```

## `summary()` function

```{r}
#|:label: summary function example

summary(ucr2017)
```

## `plot()` function

```{r}
#|label: plot function example

plot(x = ucr2017$actual_murder, y = ucr2017$actual_assault_aggravated)
```

## `View()` function

```{r}
#|label: view function example

View(ucr2017)
```

```{r}
#|label: testing to view only pre-defined number of rows

View(head(ucr2017))
```
