---
title: "Testing your code"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/07/14
---

-   Testing functions

-   Testing data collections

# 8.1 Why test your code?

-   Mistakes:

    -   With coding - those that prevent code from working (i.e give you an error message and don't run the code)

    -   Those that run the code but give you the wrong result.

```{r}
#|label: example data

example_data <- data.frame(
  year = c(2000, 2000, 2001, 2001),
  crime_type = c("murder", "theft", "murder", "theft"),
  crime_count = c(100, 100, 200, 50))

example_data
```

```{r}
#|label: example of a mistake with coding

# example_data[example_data$crime_type = "murder", c("year", "crime_count")]
# 
# Error: unexpected '=' in "example_data[example_data$crime_type ="
```

```{r}
#|label: example of a mistake with getting the wrong result

example_data[example_data$crime_type != "murder", c("year", "crime_count")]
```

# 8.2 Unit tests

-   A unit test is simply a conditional statement where you have some input, usually a function with some parameters set, and state what you expect the result to be.

```{r}
#|label: load the testthat package

library(testthat)
```

```{r}
#|label: example usage of the expect_equal() function

add_2 <- function(number) { return(number + 2) }

expect_equal(add_2(2), 4)
```

```{r}
#|label: sample test fail

expect_equal(add_2(2), 5)
```

-   Three ways to run the unit tests:

    -   **Method 1: Manual execution** Run each test line directly like a normal R script. This is inefficient and doesn't leverage RStudio's testing features, but works with regular R projects (no package required).

    -   **Method 2: Using `test_dir()`** Use the `testthat` package's `test_dir()` function, typically written as `test_dir(test_path())` to run all test files in the correct folder. This provides a summary showing passed, failed, warned, and skipped tests. You must first load all functions and data that the tests reference, either manually or using Ctrl+Shift+L (Cmd+Shift+L on Mac) to load all project files.

    -   **Method 3: Keyboard shortcut** Use Ctrl+Shift+T (Cmd+Shift+T on Mac) to automatically load all files and run all tests in one step. This is the quickest method but only works with R Packages, not regular R Projects.

## 8.2.1 Modular test scripts

```{r}
#|label: load the usethis package

library(usethis)
```

```{r}


use_test("Test")
```

## 8.2.2 How to write unit tests

-   **Basic Test Structure:**

    -   Individual tests use `expect_equal()` format: `expect_equal(2 * 2, 4)` checks if 2 \* 2 equals 4

    -   Tests are organized within `test_that()` groups: `test_that("multiplication works", {})` with tests inside the curly braces

    -   Groups help organize related tests and make it easier to identify where failures occur

-   **Running Tests:**

    -   Run individual `expect_equal()` lines (requires `library(testthat)` first)

    -   Run entire `test_that()` groups by highlighting all code or just the lines with curly braces

    -   Failed tests will specify which group failed based on the group name

-   **Real-World Example:** The author provides an example from testing US Border Patrol data scraping:

    -   Test file loads the data to be tested at the beginning

    -   Uses multiple `test_that()` groups with `expect_equal()` tests inside each

    -   Tests check if scraped PDF data matches expected values determined by manually examining the PDFs

    -   Example test: checking if "cocaine_pounds" column values for "coastal border" sectors equal a specific vector of numbers

    -   For large datasets, you can test a subset of values rather than every single value

-   **Key Principle:** Tests verify that specific inputs produce expected outputs, helping ensure code works correctly, especially for data processing tasks like PDF scraping.

## 8.2.3 What to test

-   **What to Test:**

    -   **Every parameter** in the function with various inputs

    -   **Edge cases and incorrect usage** - think of ways the function could be misused and test for those scenarios

    -   **Different input types** - for an `add_2()` function, test what happens with strings ("2") instead of numbers, or vectors instead of single numbers

    -   **Comprehensive coverage** - be thorough and cover all your bases

-   **Efficiency Considerations:**

    -   **Avoid redundant tests** - if you've tested `add_2(2)` equals 4, don't also test `add_2(3)` equals 5 since they test the same functionality

    -   **Consider your audience** - if you know users won't input strings to a numeric function, testing for that may be unnecessary

    -   **Time investment** - unit tests are highly time-efficient for preventing costly mistakes, but don't over-invest in writing them

-   **General Philosophy:**

    -   **Better to have more tests than fewer**, especially when starting out

    -   **Experience teaches balance** - you'll learn what's worth testing over time

    -   **Focus on value** - consider whether writing a particular test is a good use of your time

### 8.2.3.1 Tests for research projects

-   **Limited Testing Opportunities in Research:**

    -   Research typically involves taking existing data, cleaning it, and running regressions

    -   Few opportunities for traditional unit testing in this workflow

    -   **Best checks are often domain knowledge** - examining results to see if they make sense and align with prior literature

-   **When to Test Research Data:** Two main scenarios where data testing is valuable:

    -   **Data scraping verification** - When you're scraping data (web or PDF), tests can verify the scraping worked correctly

    -   **Changing datasets** - When you expect data to be updated and want to ensure consistency across versions

-   **Practical Examples:**

    -   Test that only your target state appears in the data (catches cases where new data releases include unwanted states)

    -   Check that broad categories remain consistent even as exact values change

    -   Verify expected groups/categories are still included after data updates

-   **When NOT to Write Tests:**

    -   **Static datasets** - If you're using a specific dataset that never changes, regular code checks in your main script are more appropriate than dedicated tests

    -   **One-time analyses** - Better to use ad-hoc verification code rather than formal test infrastructure

-   **Key Principle:** Unit tests for research data are most valuable when you need to catch systematic errors in data processing or ensure consistency across changing datasets, rather than for typical one-time data analysis workflows.

### 8.2.3.2 Tests for data collection

-   **Key Difference from Normal Testing:**

    -   **Normal tests:** Avoid redundancy - if you test that 2 + 2 = 4, don't test 2 + 3 = 5

    -   **Data collection tests:** Be far more thorough and test multiple similar scenarios

    -   **Why More Thorough Testing is Needed:** Small differences in source data can break code at different points. Example: A 10-page PDF where page 1 has 5 columns but pages 2-10 have 6 columns - testing only page 1 would miss the failure on other pages.

-   **Common Problems:**

    -   Parts of data are "weird" and cause incorrect scraping

    -   Code often fails silently without obvious error messages

    -   Inconsistencies across different sections of the same source

-   **Testing Strategy:**

    -   **Random sampling:** Test a few random pages/sections and random rows/columns for good coverage

    -   **Targeted testing:** Look for atypical parts of the PDF/website and test those specifically

    -   **Scale appropriately:** More tests for longer/more complex sources

-   **Balance Considerations:**

    -   **Better to over-test than under-test** for data collection

    -   **Diminishing returns:** Don't spend so much time testing that you can't work on actual code

    -   **Experience-dependent:** Learning the right balance comes with practice

    -   **No hard rules:** What constitutes "too much" or "too little" testing varies by situation

The core message is that data collection code requires more comprehensive testing due to the unpredictable nature of source data variations.

# 8.3 Test-driven development (TDD)

-   **What is TDD:**

    -   Write tests **first**, then write code to meet those tests

    -   Extension of project planning principles from Chapter 5 (writing out steps, sketching figures/tables)

    -   Involves writing tests for all intended functions, their parameter variations, and data collection/cleaning processes

-   **Benefits of TDD:**

    -   **Forces deep thinking** about function design and interactions

    -   **Identifies potential issues early** - the author often realized planned approaches wouldn't work while writing tests

    -   **Prevents problems** before coding begins

    -   **Improves project planning** and code architecture

-   **Why It's Advanced:**

    -   Requires knowing **exactly (or mostly) what you need to do**

    -   Need to anticipate **likely problems** each approach will face

    -   Demands significant experience to implement effectively

-   **Recommendation:** **Wait until you're experienced** with R or programming generally before attempting TDD

-   **Key Insight:** TDD is essentially a sophisticated planning tool that helps you think through your entire approach before writing any actual code, but it requires substantial programming experience to use effectively. It's more about preventing problems through careful upfront design than just testing code after it's written.
