---
title: "Geocoding"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/08/30
---

-   Data:

    -   san_francisco_active_marijuana_retailers.csv

## 24.1 Geocoding a single address

-   Geocoding is the process of taking an address (e.g. 123 Main Street, Somewhere, CA, 12345) and getting the longitude and latitude coordinates of that address.

## 24.1 Geocoding a single address

```{r}
#|label: load the tidygeocoder package

library(tidygeocoder)
```

```{r}
#|label: example of geocode() function - function expects dataframe, not tibble

# geocode("750 Race St. Philadelphia, PA 19106")

# Error: .tbl is not a dataframe. See ?geocode
```

```{r}
#|label: corrected example

address_to_geocode <- data.frame(address = 
                                   "750 Race St. Philadelphia, PA 19106")

geocode(address_to_geocode, address)
```

```{r}
#|label: address column can be used with or withoud quotes

geocode(address_to_geocode, address = "address")
```

```{r}
#|label: method parameter to geocode function - Using Open Street Map as a source

example <- geocode(address_to_geocode,
                   address = "address",
                   method = "osm")
example
```

```{r}
#|label: method parameter to geocode function - Using US Census Bureau’s geocoder as a source

example <- geocode(address_to_geocode, 
                   address = "address",
                   method = "census")
example
```

```{r}
#|label: method parameter to geocode function - Using ArcGIS geocoder as a source

example <- geocode(address_to_geocode,
                   address = "address",
                   method = "arcgis")
example
```

```{r}
#|label: convert the result to tibble

example <- geocode(address_to_geocode,
                   address = "address",
                   method = "arcgis")
example <- data.frame(example)
example
```

```{r}
#|label: method parameter to geocode function - Using Open Street Map as a source with full results

example <- geocode(address_to_geocode,
                   address = "address",
                   method = "osm",
                   full_results = TRUE)
example <- data.frame(example)
example
```

For OSM as a source we also get information:

-   about the address, such as what type of place it is,

-   a bounding box which is a geographic area right around this coordinate,

-   the address for those coordinates in the OSM database, and a bunch of other variables that don’t seem very useful for our purposes such as the “importance” of the address.

```{r}
#|label: method parameter to geocode function - Using US Census Bureau’s geocoder as a source with full results

example <- geocode(address_to_geocode,
                   address = "address",
                   method = "census",
                   full_results = TRUE)
example <- data.frame(example)
example
```

-   The Census results are similar to the OSM results and also have the matched address to compare your inputted address to.

-   Most of the columns are just the address broken into different pieces (street, city, state, etc.) so are mostly repeating the address again in multiple columns.

```{r}
#|label: method parameter to geocode function - Using ArcGIS geocoder as a source with full results

example <- geocode(address_to_geocode,
                   address = "address",
                   method = "arcgis",
                   full_results = TRUE)
example <- data.frame(example)
example
```

-   For the ArcGIS results we have the matched address again, and then an important variable called “score,” which is basically a measure of how confident ArcGIS is that it matched the right address.

-   Higher values are more confident, but in my experience anything under 90-95 confidence is an incorrect address.

-   These results also repeat the longitude and latitude columns as “location.x” and “location.y” columns, and I’m not sure why they do so.

## 24.2 Geocoding San Francisco marijuana dispensary locations

```{r}
#|label: load the readr packae

library(readr)
```

```{r}
#|label: read data into R

marijuana <- read_csv("data/san_francisco_active_marijuana_retailers.csv")
marijuana <- data.frame(marijuana)
```

```{r}
#|label: check the first 6 rows

head(marijuana)
```

```{r}
#|label: county is always San Francisco - remove that part from address

marijuana$Premise.Address <- gsub(" County: SAN FRANCISCO",
                                  "", marijuana$Premise.Address)
```

```{r}
#|label: check the first 6 rows of the column

head(marijuana$Premise.Address)
```

```{r}
#|label: geocode the Premise Address column

marijuana <- geocode(marijuana, "Premise.Address")
```

```{r}
#|label: check the summary of longitude column

summary(marijuana$long)
```

```{r}
#|label: check the summary of latitude column

summary(marijuana$lat)
```

```{r}
#|label: remove longitude and latitude columns for replacement

marijuana$long <- NULL
marijuana$lat <- NULL
```

```{r}
#|label: Using ArcGIS geocoder as a source for San Francisco marijuana data

marijuana <- geocode(marijuana,
                     address = "Premise.Address",
                     method = "arcgis")
```

```{r}
#|label: check longitude again

summary(marijuana$long)
```

```{r}
#|label: check latitude again

summary(marijuana$lat)
```

-   No more NAs, which means that we successfully geocoded our addresses.

-   Another check is to make a simple scatterplot of the data.

-   Since all of the data is from San Francisco, they should be relatively close to each other. If there are dots far from the rest, that is probably a geocoding issue.

```{r}
#|label: create a scatter plot to check longitude and latitude values

plot(marijuana$long, marijuana$lat)
```
