---
title: "Reading and writing data"
author: "Mardan Mirzaguliyev"
format: html
editor: visual
date: 2025/06/30
---

## 4.1 Reading data into R

### 4.1.1 R

#### 4.1.1.1 .rda and .rdata files

```{r}
#|label: reading .rda file

load("data/sf_neighborhoods_suicide.rda")
```

#### 4.1.1.2 .rds files

```{r}
#|label: reading .rds file

rds_example <- readRDS("data/shr_1976_2020.rds")

head(rds_example)
```

### 4.1.2 Excel

```{r}
#|label: load the readr package

library(readr)
```

```{r}
#|label: reading .csv file

shootings <- read_csv("data/fatal-police-shootings-data.csv")

head(shootings)
```

```{r}
#|label: convert a tibble to a data frame

shootings <- as.data.frame(shootings)
```

```{r}
#|label: load the readxl function

library(readxl)
```

```{r}
#|label: reading .xls and .xlsx files

# Stop, question and frisk
sqf <- read_excel("data/sqf-2019.xlsx")

head(sqf)
```

### 4.1.3 Stata

```{r}
#|label: load the haven package

library(haven)
```

-   **`read_dta()`** - Stata file, extension ".dta"

-   **`read_sas()`** - SAS file, extension ".sas"

-   **`read_sav`** - SPSS file, extension ".sav"

```{r}
#|label: reading .dta file

shootings <- read_dta("data/fatal-police-shootings-data.dta")

head(shootings)
```

### 4.1.4 SAS

```{r}
#|label: reading .sas file

shootings <- read_sas("data/fatal-police-shootings-data.sas")

head(shootings)
```

### 4.1.5 SPSS

```{r}
#|label: reading .sav file

shootings <- read_sav("data/fatal-police-shootings-data.sav")

head(shootings)
```

### 4.1.6 Fixed-width ASCII

-   ASCII stands for American Standard Code for Information Interchange. It's a fundamental system that assigns unique numerical values to characters like letters, numbers, and symbols. Think of it as a translator, allowing computers to understand and display text. The original ASCII used a 7-bit system, supporting 128 characters. While still foundational, it's largely been replaced by Unicode (like UTF-8), which can handle characters from almost all global languages while remaining compatible with ASCII. (Gemini (2025-07-01))

-   A fixed-width ASCII file is a text file where data is organized into columns with a predefined, unchanging number of characters for each column on every line. There are no delimiters; column separation is determined solely by their fixed positions and lengths. Excess space is filled with padding (usually spaces). These files require a separate "layout" document to understand the data structure. (Gemini (2025-07-01))

```{r}
#|label: load the asciiSetupReader package

library(asciiSetupReader)
```

```{r}
#|label: reading a fixed-width ASCI file

shr <- read_ascii_setup(data = "data/2020_SHR_NATIONAL_MASTER_FILE.txt",
                        setup_file = "data/ucr_shr.sps")

head(shr)
```

## 4.2 Writing data

-   `save()` - R file, extension ".rda" and ".rdata" (base R)

-   `saveRDS()` - R file, extension ".rds" (base R)

-   **`write_csv()`** - Excel file, extension “.csv” (`readxl` package)

-   **`write_dta()`** - Stata file, extension “.dta” (`haven` package)

-   **`write_sas()`** - SAS file, extension “.sas” (`haven` package)

-   **`write_sav()`** - SPSS file, extension “.sav” (`haven` package)

### 4.2.1 R

#### 4.2.1.1 .rda and .rdata

```{r}
#|label: writing .rda and .rdata files

save(shootings, file = "data/shootings.rda")
```

#### 4.2.1.2 .rds

```{r}
#|label: writing .rds file

saveRDS(shootings, "data/shootings.rds")
```

### 4.2.2 Excel

```{r}
#|label: writing .csv file

write_csv(shootings, "data/shootings.csv")
```

### 4.2.3 Stata

```{r}
#|label: writing .dta file

write_dta(shootings, "data/shootings.dta")
```

### 4.2.4 SAS

```{r}
#|label: writing .sas file

write_xpt(shootings, "data/shootings.sas")

# Warning: `write_sas()` was deprecated in haven 2.5.2.
# ℹ Please use `write_xpt()` instead.
```

### 4.2.5 SPSS

```{r}
#|label: writing .sav file

write_sav(shootings,"data/shootings.sav")
```
